---
title: "Lab2_Sp21"
author: "sbsambado"
date: "1/16/2021"
output: html_document
---

Lab 2: Introduction to R Markdown and Data Analysis

# 3 common syntax errors
**spelling** 
+install.packages('readr'), correct
+instal.package('readr') , not correct

**capitilization**
+ install.packages('readr'), correct
+ Install.Packages('readr'), not correct

**punctuation**
+ install.packages('readr'), correct
+ instll_packages('readr'), not correct


# R basics

variable <- function(argument)


#List of useful functions to create and explore data structure
* Basic mathematical functions for summary statistics
    + `sum()`
    + `mean()`
    + `min()`
    + `max()`
    + `sd()`
    + `median()`
    + `summary()`
    
* Functions to create & explore data structure 
    + `c()` - concatenate function
    + `:` - colon operator for sequence generation
    + `seq()` - sequence generation
    + `rep()` - replicate elements of vectors and lists
    + `View()` - invoke a data viewer
    + `length()` - length of an object
    + `class()` - object class
    + `head()` - return the first 6 rows of an object
    + `tail()` - return the last last 6 rows an object
    + `dim()` - returns the dimensions of an object 
    + `nrow()` - number of rows
    + `ncol()` - number of columns
    + `str()` - display the structure of each column of an object
    + `names()` - display the names of an object, typically the column names
    + `nlevels()` - categorial levels within an object
    + `cbind()` - combine objects by columns
    + `rbind()` - combine objects by rows



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#install.packages('readr') # if first time using readr
## hashtag is making a comment, it is not seen as code
# annotating your code is great practice! 

#install.packages("readr")
library(readr)
library(ggplot2)
library(tidyverse)
library(knitr)


library(ggplot2)
```


#Exercises

```{r}
# A
bat_tongues <- read_csv("battongues.csv")
#View(bat_tongues)

#C
bat_copy <- bat_tongues

names(bat_copy)


# let's change them
names(bat_copy) <- c("spp", "palate", "tongue")
names(bat_copy)

# mean of tongue
mean(bat_copy$tongue)


# other stats
summary(bat_copy$palate)

#### Caffeiene 

caffeine <- read_csv("caffeine.csv")
names(caffeine) <- c("brand", "caffeine_mg")


# basic stats
var(caffeine$caffeine_mg)
mean(caffeine$caffeine_mg)
max(caffeine$caffeine_mg)
min(caffeine$caffeine_mg)
sd(caffeine$caffeine_mg)
median(caffeine$caffeine_mg)
range(caffeine$caffeine_mg)
quantile(caffeine$caffeine_mg)

# or just run summary()
summary(caffeine$caffeine_mg)

# how to analyze categorical data
table(caffeine)
table(caffeine$brand)

# plot it!

# HINT: The basics of ggpploting
# ggplot(data = DATA, aes(x = INDEPENDENT VARIABLE, y = DEPENDENT VARIABLE)) +
# geom_point()+

ggplot(data = caffeine, aes(x = brand, y = caffeine_mg)) +
  geom_histogram(stat = "identity") +
  theme_minimal()+
  theme(axis.text.x.bottom =  element_text(size = 8, angle = 90))
  


```

#Homework

FOR SPRING OF 2021 RESHAPE THE HOMEWORK QUESTIONS TO GET STUDENTS THINKING MORE ABOUT STRUCTURE OF DATA VS TRYING TO FIND COMMANDS
**Question 1.**
```{r}
# load in required data
mammal <- read_csv("mammals.csv")

# let's look at file
#View(DATAFRAME)

# names(DATAFRAME) <- c("COLUMN1NAME", "COLUMN2NAME", "COLUMN3NAME")

# A. Which mammal has the largest brain size:body ratio?

  # make new variable column
  # DATAFRAME$VARIABLE <- DATAFRAME$VARIABLE1 / DATAFRAME$VARIABLE2

  # find max ratio
# DATAFRAME[which.max(DATAFRAME$VARIABLE),]


# B. Which mammal has the smallest ratio?
# DATAFRAME[which.mIN(DATAFRAME$VARIABLE),]

# C. What is the mean brain size:body ratio? Are humans above or below the mean?
# mean(DATAFRAME$VARIABLE)

# find human ratio

# Option 1 (prefered)
# DATAFRAME[DATAFRAME$VARIABLE == "OBJECT",]

# Option 2
# DATAFRAME[DATAFRAME$VARIABLE == "OBJECT", "SPECIFICVALUE_OF_OBJECT"]


```



**Question 2**
```{r}
# load data

birds <- read_csv("desert_birds.csv")
#View(birds)

# A. How many desert bird species are rare (<= 10 counts)

# make subset

# VARIABLE_NAME <- subset(DATAFRAME, VARIABLE <= 10)
#find how many rows there are in subset

# Option 1. Look at length of vector

#length(DATAFRAME$VARIABLE)
#dim(DATAFRAME)
#nrow(DATAFRAME)

# Option 2. Look at dataset summary
#print(DATAFRAME)
#summary(DATAFRAME)

# This is different than finding the total sum of birds within these rare species
#sum(DATAFRAME$VARIABLE)
```

**Question 3**
```{r}

# load data
endangered <- read_csv("endangered.csv")

#View(endangered)

# A. What proportion of endangered species are vertebrates?

# Option 1

# VARIABLE_NAME <- subset(DATAFRAME, VARIABLE == "yes")
# sum(DATAFRAME1$VARIABLE) / sum(DATAFRAME2$VARIABLE)



```
