---
title: "Lab 2"
author: "sbsambado"
date: "3/23/2020"
output: html_document
---

EEMB 247 Computer lab 2: More useful things in R, and linear Dynamical systems in continuous time and discrete time

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

A few more useful things in R
```{r}
# Logical operators (TRUE, FALSE, NA)

x = 11:18
x
y = x < 15
y

# <= : less than, or equal
# >= :greater than, or equal
# != : inequality

# if y and z are logical expressions, then y & z is their intersection 'and', y | z is their union 'or', and !y is the negation of y

x = 1:10
y = x < 5
z = x < 7

z
y & z
y | z

## Missing values (NA)

# function is.na(x) gives a logical vector of the same size as x with value TRUE if and only if the corresponding element in x is NA

z = c(1:3, NA)
z

ind = is.na(z)
ind

# x == NA is different than is.na(x) since NA is a marker for a quantity that is not available

x = c(1:10, NA)
x

# x == NA

# the second 'missing' values are produced by numerical computation, so-called Not a Number 'NaN' (e.g. 0/0, Inf - Inf)

## Index Vectors; selecting and modifying subsets of a data set
z = c(1:3, NA)
z

y = z[!is.na(z)] # removes NA
y

w = (z + 1)[(!is.na(z)) & z >1] # places values of vector z+1 for which the corresponding value in z was both non-missing and greater than 1
w

# index vectors can be a vector of positive integers
x = seq(12, 30)
x
x[6] # selects 6th component of x
x[1:10] # selects first 10 elements of x

# index vectors can be vector of negative integers. These index vector specifies the values to be excluded rather than included.
y = x[-(1:5)]
y

# index vectos can be a vector of character strings. Only applies where an object has a 'names' attribute to identify its components. In this case a sub-vector of the names vector may be used in the same way as positive integral labels
fruit = c(5,10,1,20)
names(fruit) = c('orange', 'bannana', 'apple', 'peach')
lunch = fruit[c('apple', 'orange')]
lunch

# indexed expression can also appear on the receiving end of an assignment, in which case the assignment operation is performed only on those elements of the vector

# replace any missing values in x by zeros
x = c(1,4, NA, 5, 7, NA, 12)
x

x[is.na(x)] = 0
x

# y[y<0] <- -y[y<0] has same effect as: y <- abs(y)
y = c(-4, 4, 2, -3)
y

y[y < 0] = -y[y<0]
y


## Conditional expressions
# conditional expressions allow you to do different things based on the value of a variable

# basic format
# if (condition) {
# expression(s) if true
# } else {
# expression(s) to do otherwise
#}

x = 5
if (x > 3) y = 12
y

# or

x <- 5
if (x > 3) {
  y = 12
  z = 3
} else {
  y = 0
  z = 200
}
y
```
Repetitive execution: for loops and avoiding for loops
```{r}
# a for loop allows you to loop over values in a vector or list of numbers, and perform some operations over and over again
# R has a family of functions, the 'apply' family, which can be used to avoid for loops
  # apply - apply over the margins of an array (rows/columns of a matrix)
  # lapply - apply over an object and return list
  # sapply - apply over an object and return a simplified object (an array)
  # vapply - similar to sapplym but yo specify the type of object returned by the iterations
    
    # array = homogeneous elements ; matrix = simple row and column
# apply function acts on arrays example
# apply(X, margin, FUN, ...)
  # X is array or matrix
  # margin is a variable defining how the function is applied
      # MARGIN = 1 ; applies over rows
      # MARGIN = 2 ; applies over columns
      # MARGIN = c(1,2) applies to rows and columns
  # FUN is the function you want to apply to the data

X = array(1:20, dim = c(4,5)) # 4 x 5 matrix
X

apply(X, 2, sum) # sum of columns in matrix, X
apply(X, 1, mean) # mean of rows in matrix, X

# loops in R are slow
  # 1. don't use a loop when a vectorized alternative exists
  # 2. don't grow objects (via c, cbind, etc.) during the loop
  # 3. Allocate an object to hold the results and fill it in during loop

## BASIC STRUCTURE OF FOR LOOP ##

# for (variable in vector) {
#   expression(s)
# }

# want to add 5 to each element in vector x and store it in y
x = c(1,3,2,4,5,7,9,0,3)
y = rep(0, length(x))

for (i in 1:length(x)) {
  y[i] = x[i] + 5
}
y

# however, this could be expediated
y = x +5
y

# allocate an object y to hold the answer
y = rep(0, length(x) - 1)

for (i in length(x) - 1) {
  y[i] = x[i] + x[i+1]
}
y
```

Numerical solutions to discrete time models using for loops
```{r}
# for discrete time equations, we can obtain numerical solutions to discrete-time difference equations using for loops

R = 2.5
initial_N = 0.3
max_time = 20

N = rep(0, max_time + 1)
N[1] = initial_N
for (t in 1:max_time) {
  N[t+1] = R*N[t]*(1 - N[t])
}

times = seq(0, max_time)
plot(times, N, type = 'b', xlab = 'time', ylab = 'N')

```

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
