---
title: "Lab 3 - Equilibria, Stability"
author: "sbsambado"
date: "3/23/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Data file from class:

library(readr)
frog <- read_csv("~/Desktop/frog_data.csv")
View(frog)


Computer lab 3: Calculating equilibria, plus R tutorial on inputting and manipulating data
```{r}
#library(deSolve)
#library(rootSolve)
#library(ggplot2)
```

Calculating equilibrium densities, analytically and numerically

Determine equilibrium values of the state variables for models consisting of systems of ODEs

An equilibrium is a constant solution to a set of differential equations

The equilibrium is the value of the state variables in a model for which the rates of change of the state variables are equal to 0

mathematically, for differential equation dN/dt = f(N), the equilibrium value of state variable N, which we denote as N*, is the value of N* for which f(N*) = 0

# 3 Methods for determining the equilibrium solutions to models ##

   1. Simulation: for specified values of parameters and initial conditions of the state variable, we can simply run the model for a long time period and see if the state vairbales level off at an equilibrium value
   pros : it's easy, we can attempt this with any model
   cons : requires specifying values of parameters/state variables; trajectory will level off only if equilibrium is stable; equilibrium may take a long time to get there; some models can have multiple equilibria

   2. Analytically: For some models, we can write down an analytical expression for N* by using ALGEBRA to solve equation f(N*) = 0 for N*
   cons: sometimes it's not possible to obtain an explicit, closed-form solutions to the equations or involve messy algebra

   3. Numerically: We can use one of a number of numerical methods to numerically find solution to N* to the equation f(N*) = 0 for specified values of parameters
   
A simple predator-prey model
```{r}
# Prey alone
# what is the equilibrium density for the prey population in the absence of the predator?

# solve for the equilibrium numerically
I_N = 10
dN = 0.2

RHS_prey = function(N, I_N, dN) {
  I_N - dN*N
}

# find equilibrium using 'uniroot.all', specifying the interval of values of N to search over: uniroot.all(function, interval, ...)

prey_equilibrium = uniroot.all(RHS_prey, c(0, 100), I_N = I_N, dN = dN)
prey_equilibrium

# plot how prey density changes as function of the prey immigration rate
dN = 0.2
I_N_all = seq(0, 20, 0.1)

#set up a vector to store prey equilibrium values
prey_equilibrium = rep(0, length(I_N_all))

for (i in 1:length(I_N_all)) {
  # calculate the equilibrium value of N for each value of I_N
  prey_equilibrium[i] = uniroot.all(RHS_prey, c(0, 200), I_N = I_N_all[i], dN = dN)
}

plot(I_N_all, prey_equilibrium, type = 'l', col = 'blue', xlab = 'prey immigration rate', ylab = 'prey equilibrium, N*')

# OR avoid for loop using sapply function

dN = 0.2
I_N_all = seq(0, 20, 0.1)

myfun = function(x) {
  uniroot.all(RHS_prey, c(0, 200), I_N = x, dN = dN)
}

prey_equilibrium=sapply(I_N_all, myfun)

# Equilibrium with both predator and prey

# Solve for equilibrium numerically
# multiroot to solve two equations simultaneously
  # given a vector of n variables, and a set of n (nonlinear) equations in these varaibles, 'multiroot' estimates the root of the equations (i.e. the values of the variables where all function values = 0)
  # syntax for calling multiroot: multiroot(f, start, ...)
    # f is the function for which the root is sought; must return a vector with as many values as the length of start

RHS_prey_pred = function(x, parms) {
  N = x[1]
  P = x[2]
  I_N = parms[1]
  dN = parms[2]
  a = parms[3]
  c = parms[4]
  dP = parms[5]
  
  F1 = I_N - dN*N - a*N*P
  F2 = c*a*N*P - dP*P
  
  return(c(F1, F2))
}

parameters = c(I_N = 10, dN = 0.2, a = 0.01, c = 0.5, dP = 0.1)
x = c(N = 10, P = 10)

#call multiroot
eqm = multiroot(RHS_prey_pred, x, parms = parameters)

# equilibrium values are stored in the root part of eqm
eqm$root

# will tell you the precision, and the number of iterations in numerical routine took
eqm
```

Calculating equilibrium values in discrete time models
```{r}
# Using Nicholson-Bailey host-parasitoid model

# Solve for the equilibrium numerically
R = 2
a = 0.05
c = 0.9

parameters = c(R = R, a = a, c = c)

# This function returns the set of equations that are to be set equal to 0
Nicholson_Bailey_eqm = function(x, parms) {
  H = x[1]
  P = x[2]
  R = parms[1]
  a = parms[2]
  c = parms[3]
  
  F1 = H - R*H*exp(-a*P)
  F2 = P - c*H*(1 - exp(-a*P))
  
  return(c(F1, F2))
}

x = c(H = 10, P = 10)

eqm = multiroot(Nicholson_Bailey_eqm, x, parms = parameters)

eqm$root

# check to see if it found the correct equilibrium
Pstar = log(R)/a
Hstar = Pstar*R/ (c*(R - 1))

Pstar
Hstar
```
R tutorial, Part 3: Importing and Manipulating data in R
```{r}
getwd()
library(readr)
frog <- read.csv("~/Desktop/frog_data.csv")
View(frog)

# examine data structure 
class(frog)

# data frames are good because you can have a cotinuous variable (i.e. weight) and a categorical variable (e.g. color)
# useful dataframe functions
  # head()
  # tail()
  # dim()
  # nrow()
  # ncol()
  # str() # structure of each column
  # names() # either do colnames(frog) or names(frog)

# look at fifth column
colnames(frog)[5]
frog[,5]
class(frog[,5])

str(frog)
  

# 5.1 Addressing data
  # 3 main ways to address data in a data frame
    # 1. By index
    # 2. By logical vector
    # 3. By name (columns only)

# By index. We can see dimensions, shape, of data frame
dim(frog)
# want to get a single value from data frame, we can provide an index in [] 
x = frog[1,2] # value in 1st row and 2 column of frog and store it in x
x

# Select whole sections
frog[1:10, 1:4] # select first 10 frogs (rows), first 4 variables (columns)

# use 'c' to select non-continguous values
frog[c(1,8,10,12), c(2,3,6)]

# select all columns from row 5
frog[5,]

# select all rows from column 4
frog[,4]

# 5.2 Logical addressing
# logical vectors can be created using relational operators (e.g. >,<, ==, !=)
x = frog[,3] > 10
x

# 5.3 By name
names(frog)

# columns can be addressed using the $ operator
frog$sex
frog$virus

# OR
frog['sex']
frog['virus']

class(frog$sex) # returns a vector
class(frog['sex']) # returns a dataframe

# a subsection of a dataframe is called a slice. We can take slices of character vectors as well
element = c('o', 'x', 'y', 'g', 'e', 'n')
element[1:3]

element[1:4]
element[4:1]

element[-1] # removes first element
element[-1:-4] # removes 1:4

# select all male frogs 
index <- frog$sex == 'M'
frog[index,]

# combining indexing and assignment 
x = c(5,3,7,10,15,13,17)

# when x is greater than 10 set it to 0
x[x > 10] = 0
x

# combine indexing and assignment to 'correct' the weight column in the frog so that all values weight that are greater than 12 are set to equal 12
index <- frog$weight > 12
frog[index,]
frog[index,3] <- 12

# OR

frog$weight[index] <- 12

## 6 Factors

# factors are used to repreesnt categorical data. Factors can be ordered or unordered
# factors are stored as integers, and have labels associated with thse unique integer. While factors look like character vectors, they are actually integers
# once created, factors can only contain a pre-defined set values, known as levels

sex = factor(c('male', 'female', 'female', 'male'))
sex

# R will assign 1 to level 'female' and 2 to 'male' because f comes before m
# check number of levels using nlevels()
levels(sex)
nlevels(sex)

# want to specify the order
food = factor(c('low', 'high', 'medium', 'high', 'low', 'medium', 'high'))
levels(food)

food = factor(food, levels = c('low', 'medium', 'high'))
levels(food)

food = factor(food, levels = c('low', 'medium', 'high'), ordered = TRUE)
levels(food)
min(food) # now in R's memory as factors represented by numbers (1,2,3)

# converting factors to a character vector
as.character(food)

# converting factor to numeric vector
f = factor(c(1,5,10,2))
as.numeric(levels(f))[f]


# function 'table()' tabulates observations and can be used to create bar plots quickly
exprmt = factor(c('treat1', 'treat2', 'treat1', 'treat3', 'treat1', 'control', 'treat1', 'treat2', 'treat3'))
table(exprmt)

barplot(table(exprmt))

# reorder barplot
exprmt = ordered(exprmt, levels = c('treat1', 'treat2', 'treat3', 'control'))
barplot(table(exprmt))
# OR
exprmt = factor(exprmt, levels = c('treat1', 'treat2', 'treat3', 'control'))
barplot(table(exprmt))

## removing levels from a factor
# create a new dataset with only blue and green frogs from our original frog dataset

bgfrog = frog[(frog$color == 'blue') | (frog$color == 'green'),]
levels(bgfrog$color) # R still thinks 'red/yellow' are valid levels

# remove any unused levels of factors
bgfrog = droplevels(bgfrog)
levels(bgfrog$color)
```

Manipulating Data
```{r}
head(frog)


# select data we want to create a new temporary data frame, and then perform the calculation on this subset
gfrog = frog[frog$color == 'green',]
max(gfrog$worms)

# combine the selection and the function call
max(frog$worms[frog$color == 'green'])

# common calculations

min(frog[,3]) # min weight of frog
# OR
min(frog$weight)

mean(frog$worms)
median(frog$worms)
sd(frog$worms) # sd of # of worms per frog

## using 'summary()' will give us summary statistics of all columns in dataset
summary(frog)

## 'apply' allows us to repeat a function on all of rows (MARGIN = 1) or columns (MARGIN = 2)

# combine the two types of parasites (worms and fungus) and ifnd total number of parasites per frog
parasites = apply(frog[,c(5,7)], 1, sum)
parasites
# OR
apply(frog[,c(5,7)], MARGIN = 1, sum)

# or create a new column in dataset that is the sum of the number
frog$parasites = frog$worms + frog$fungus
frog$parasites

# or calculate row-wise or column-wise means with 'rowMeans' and 'colMeans'

## Outputting data
write.table(frog, file = 'newfrog.csv')
```


```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
