---
title: "Lab 8 - Stage structured and spatially structured models"
author: "sbsambado"
date: "3/25/2020"
output: html_document
---
EEMB247 Computer lab 8: Spatially strucutred and stage-structured models

Consider a system with two developmental stages, modeled in continuous time. This could represent a dynamical system in which the entities in the system transition from one state to another

If we model this system of ODEs, in which individuals mature at a constant per individual rate of mj, then the distribution of times psent in J stage will be exponentially-dsitributed, teh average time spent in J stage will be Tj = 1/mj
```{r}

require(deSolve)

twostage <- function(t, y, parms) {
  J = y[1]
  A = y[2]
  mJ = parms[1]
  
  dJdt = -mJ*J
  dAdt = mJ*J
  
  ODEs = c(dJdt, dAdt)
  list(ODEs)
}

mJ = 1/10
pars = c(mJ = mJ)

init = c(J = 100, A = 0)

tInt = seq(0,50, by = 0.1)

Sout = as.data.frame(lsoda(init, times = tInt, func = twostage, parms = pars))
mat.rate = mJ*Sout$J
plot(Sout$time, Sout$J, ylim = c(0, 100), type = 'l', col = 'black', lty = 2, lwd = 2,
     xlab = 'time', ylab = 'densities, and maturation rates')
lines(Sout$time, Sout$A, lty = 3, lwd = 2, col = 'blue')
lines(Sout$time, mat.rate, col = 'red')
legend('right', legend = c('J', 'A', 'mat.rate'), lty = c(2,3,1), lwd = c(2,2,1),
       col = c('black','blue','red'))
```

Divide juvenile classes into n = 2 identical substages
```{r}
LinChain <- function(t, y, parms) {
  Nstates = parms[1];
  m = parms[2];
  
  dJdt = rep(0, Nstates)
  dJdt[1] = - m*Nstates*y[1];
  
  if (Nstates > 1)
    {
      for(i in 2:Nstates) 
        {
           dJdt[i] = m*Nstates*y[i - 1] - m*Nstates*y[i];
    }
  }
      dAdt = m*Nstates*y[Nstates]
      
      ODEs <- c(dJdt, dAdt)
      list(ODEs)
}

m = 1.0
Nstates = 5

pars <- c(Nstates = Nstates, m = m)
init <- c(100, rep(0,Nstates))
tInt <- seq(0, 10, by = 0.1)

Sout <- lsoda(init, times = tInt, func = LinChain, parms = pars)
Juvs <- Sout[,2:(Nstates + 1)]
if (Nstates > 1) {
  TotJuvs <- rowSums(Juvs)
} else {
  TotJuvs <- Sout[,2]
}

Adults <- Sout[,(Nstates + 2)]
maturation <- m*Nstates*Sout[,(Nstates + 1)]

par(mar = c(5, 4, 4, 5) + .1)
plot(tInt, TotJuvs, type = 'l', lty = 2, lwd = 2, xlab = 'time', ylab = 'J,A',
     ylim = c(0, 100))
lines(tInt, Adults, lty = 3, lwd = 2, col = 'blue')

par(new = TRUE)
plot(tInt, maturation, type = 'l', col = 'red', xaxt = 'n', 
     xlab = '', ylab = '')
axis(4)
mtext('mat.rate', side = 4, line = 3)

legend('right', legend = c('J', 'A', 'mat.rate'), lty = c(2,3,1), 
       col = c('black', 'blue', 'red'))

```

Stage- Structured Host-parasitoid model 

Now the system has a parasitoid (P) that attacks only the J
```{r}

LinChain <- function(t, y, parms) {
  Nstates = parms[1]
  m = parms[2]
  a = parms[3]
  dP = parms[4]
  R = parms[5]
  dA = parms[6]
  
  Adult = y[Nstates + 1]
    P = y[Nstates + 2]
    dJdt = rep(0, Nstates)
    dJdt[1] = R*Adult - m*Nstates*y[1] - a*P*y[1]
    parasitism = a*P*y[1]
    
    if (Nstates > 1) {
      for (i in 2:Nstates) {
        dJdt[i] = m*Nstates*y[i - 1] - m*Nstates*y[i] - a*P*y[i];
        parasitism = parasitism + a*P*y[i];
      }
    }
    dAdt = m*Nstates*y[Nstates] - dA*Adult
    dPdt = parasitism - dP*P
    
    ODEs <- c(dJdt, dAdt, dPdt)
    list(ODEs)
}

Nstates = 3
m = 0.1
a = 0.1
dP = 1
R = 20
dA = 1.0

pars <- c(Nstates = Nstates, m = m, a = a, dP = dP, R = R, dA = dA)

init <- c(10, rep(0, Nstates), 10)
tInt <- seq(0, 100, by = 0.1)

Sout <- lsoda(init, times = tInt, func = LinChain, parms = pars)
Juvs <- Sout[,2:(Nstates + 1)]

if(Nstates > 1) {
  TotJuvs <- rowSums(Juvs) 
} else {
  TotJuvs <- Sout[,2]
}

Adults <-Sout[,(Nstates +2)]
P = Sout[,(Nstates + 3)]
maturation <- m*Nstates*Sout[,(Nstates + 1)]

plot(tInt, TotJuvs, type = 'l', lty = 2, lwd = 2, xlab = 'time', 
     ylab = 'J, A', ylim = c(0, max(TotJuvs + 10)))
lines(tInt, Adults, lty = 1, lwd = 2, col = 'blue')
lines(tInt, P, lty = 1, lwd = 1, col = 'red')
legend('top', legend =c('J', 'A', 'P'), lty = c(2,1,1), lwd = c(2,2,1),
       col = c('black', 'blue', 'red'))

```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

