---
title: "Week 5 - Separatio of timescales"
author: "sbsambado"
date: "3/24/2020"
output: html_document
---

Computer lab 4: Local stability analysis, separation of time scales, plus R tutorial on the basics of graphing in R
```{r}
#library(deSolve)
#library(rootSolve)
```

Local stability analysis

3 steps to perform local stability analysis to determine whether an equilibrium is locally stable:
  1. Calculate the equilibrium
  2. Use Taylor's theorem to linearize the differential equation near equilibrium
  3. Solve the linear differential equations describing the dynamics of a small perturbations from equilibrium
```{r}
# Use R to determine numerically whether an equilibrium is locally stable

# rootSolve package has 'jacobian.full' function that generates the Jacobian matrix for right hand side of a set of ordinary differential equations
# from the Jacobian matrix we can determine whether or not the equilibrium is locally stable by calculating the eigenvalues of this matrix
# for continuous-time model, the equilibrium will be locally stable if the real parts of all of the eigenvalues are negative
# syntax for 'jacobian.full' : jacobian.full(y, func, parms, ...)
  # y is state variable values at which the Jacobian is estimated
  # func is function specifying the ODEs
  # parms is the parameter vector

# calculate equilibrium using multiroot

PredPrey_eqm <- function(x, parms) {
  N = x[1]
  P = x[2]
  I_N = parms[1]
  dN = parms[2]
  a = parms[3]
  c = parms[4]
  dP = parms[5]
  
  c(F1 = I_N - dN*N - a*N*P,
    F2 = c*a*N*P - dP*P)
}

# function returning the gradients for the model (used in Jacobian.full)
PredPrey <- function(t, state, parameters) {
  N = state[1]
  P = state[2]
  
  I_N = parameters[1]
  dN = parameters[2]
  a = parameters[3]
  c = parameters[4]
  dP = parameters[5]
 
  # rate of change
  dNdt = I_N - dN*N - a*N*P
  dPdt = c*a*N*P - dP*P 
  
  # return the rate of change
  list(c(dNdt, dPdt))
}

I_N = 10
dN = 0.2
a = 0.01
c = 0.5
dP = 0.1
parameters <- c(I_N, dN, a, c, dP)

initN = 10
initP = 10
x = c(N = 10, P = 10)

# call multiroot
eqm = multiroot(PredPrey_eqm, x, parms = parameters)
# equilirbium values are stored in the root part of eqm
eqm$root

# Numerically calculate the jacobian matrix at equilibrium
JAC = jacobian.full(y = c(eqm$root), func = PredPrey, parms = parameters)

# use 'eigen' to calculate the eigenvalues of this matrix
# if all real parts of the eigenvalues are negative, the equilibrium is locally stable
eigs = eigen(JAC)
eigs

# these results give us 2 eigenvalues and their corresponding right eigenvectors of the Jacobian matrix
# this case gives us real, negative eigenvalues, so small perturnations from equilibrium should return exponentially equilibrium

# the 'Re()' function returns the real part of a number, so in cases in which eigenvalues are complex numbers, we calculate the maximum of the ral parts of the eigenvalues using this command
max(Re(eigs$values))

# pred-prey example is boring because equlibrium is always positive
# Let's look at another model where the equilibrium can be positive/negative depending on choice of parameters (i.e. The Brusselator, an oscillatory chemical raction network)

# function returning the gradients for the model (used 'jacobian.full')
Brusselator = function(t, state, parameters) {
  x = state[1]
  y = state[2]
  k1 = parameters[1]
  k2 = parameters[2]
  k3 = parameters[3]
  k4 = parameters[4]
  
  dxdt = k1 - k2*x + k3*x^2*y - k4*x
  dydt = k2*x - k3*x^2*y
  
  list(c(dxdt, dydt))
  
  # specify values of parameters
  k2 = 2
  k3 = 0.5
  k4 = 1
  # let's try a value for 
  k1 = 2
  
  parameters = c(k1, k2, k3, k4)
  
  # we can calculate the equilibrium values for this system analytically
  x_eqm = k1/k4
  y_eqm = (k2*k4)/(k1*k3)
  
  # Numerically calculate the jacobian matrix at the equilibrium
  JAC = jacobian.full(y = c(x_eqm, y_eqm), func = Brusselator, parms = parameters)
  JAC
}

eigs = eigen(JAC)
eigs
max(Re(eigs$values))
```

Calculating stability boundaries, numerically
```{r}
# function returning the gradients for the model (used in jacobian.full)

max_eig_Brusselator = function(x, k2, k3, k4) {
  k1 = x
  x_eqm = k1/k4
  y_eqm = (k2*k4)/(k1*k3)
  
  A11 = k2 - k4
  A12 = k3*x_eqm^2
  A21 = -k2
  A22 = -k3*x_eqm^2
  A = matrix(c(A11, A12, A21, A22), nrow = 2, ncol = 2, byrow = TRUE)

  eigs = eigen(A)
  
  max(Re(eigs$values))
}

myroot = uniroot(max_eig_Brusselator, c(1,2), k2 = k2, k3 = k3, k4 = k4)

print("boundary for stability")

myroot$root

# we can make a 2D plot of stability boundary calculating the value of k1 dividing the stable and unstable regions for a range of values of some other parameters
# plot stability boundary on the k2 and k1

# function returning the gradients for the model (used in jacobian.full)
max_eig_Brusselator = function(x, k2, k3, k4) {
  k1 = x
  x_eqm = k1/k4
  y_eqm = (k2*k4)/(k1*k3)
  
  A11 = k2 - k4
  A12 = k3*x_eqm^2
  A21 = -k2
  A22 = -k3*x_eqm^2
  A = matrix(c(A11, A12, A21, A22), nrow = 2, ncol = 2, byrow = TRUE)

  eigs= eigen(A)
  max(Re(eigs$values))
}

k3 = 0.5
k4 = 1

k2_values = seq(1, 10, 0.01)
k1_boundary = rep(0, length(k2_values))
i = 0
for (k2 in k2_values) {
  i = i + 1
  myroot = uniroot(max_eig_Brusselator, c(0,10), k2 = k2, k3 = k3, k4 = k4)
  k1_boundary[i] = myroot$root
}

plot(k1_boundary, k2_values, type = 'l', xlab = 'k1', ylab = 'k2')
text(3.5, 3, 'stable')
text(1.5, 6, 'unstable')
```

R tutorial, part 4: THe basics of plotting in R
```{r}
#library(readr)
frog<- read.csv("~/Desktop/frog_data.csv")
plot(frog) # R will plot each variable against every other variable

# 1. Scatter plot
plot(fungus~worms, data = frog)
# very basic plot format: plot(variable1 ~ variable2, data = dataframe)
# other style to plots
  #xlab, ylab
  #pch: shape of data points 
  #cex: scale text and symbols relative to default
  # col = colors
  #.axis or .lab: control size or col of axis and labels
  # xlim, ylim
  # type = 'l' or 'b'

plot(fungus~ worms, data = frog,
     ylab = 'fungus(zoospores)', xlab = 'worms(number per frog)',
     cex = 2, cex.axis = 1.5, pch = 21,
     col = 'black', bg = 'grey',
     xlim = c(0, 20), ylim = c(0,100))


# 2. Histograms
hist(frog$weight)
# 'hist()' does not take a data argument, must specify using frog$weight
# want to adjust breaks, which accepts a vector
hist(frog$weight, breaks = c(0, 5, 10, 15, 20))

hist(frog$weight, breaks = seq(0,20,1), col = 'blue', 
     xlab = 'weight (g)', main = '')

# boxplot

boxplot(weight~color, data = frog,
        xlab = 'frog color', ylab = 'frog weight')
# R default for boxplot are 25th and 75th precentile (lower and upper limit quartiles), band near middle is alwasy 50th percentile (median)
# default boxplot{graphics} code
  # upper whisker = min(max(x), Q_3 + 1.5*IQR)
  # lower whisker = max(min(x), Q_1 - 1.5*IQR)
      # IQR = Q3 - Q1 is the box length 


# Save/Export Figures
# 1. open plotting 'device' and specify the file path
# 2. run the plot function
# 3. close the device

jpeg('Figure_1.jpg') # add path to this
plot(fungus~ worms, data = frog,
     ylab = 'fungus(zoospores)', xlab = 'worms(number per frog)',
     cex = 2, cex.axis = 1.5, pch = 21,
     col = 'black', bg = 'grey',
     xlim = c(0, 20), ylim = c(0,100))
dev.off()

```

ggplot
```{r}
# base plot
plot(fungus ~ worms, data = frog)

# ggplot
ggplot(data = frog, aes(x = worms, y = fungus)) + geom_point()

# parts of a ggplot
# 1. function 'ggplot()'
# 2. the arguments
  # data, aes, geom, stats, facets, scales, theme, and others

# data: assign their plot function to a variable
myplot <- ggplot(data = frog)

# aes: how data are visually represented, once will be independent (x), another will be dependent (y) 
myplot <- ggplot(data = frog, aes(x = worms, y = fungus))

# geom: how we create the layer we actually see as our figure
  # geom_point(), geom_line(), geom_boxplot(), geom_text()
myplot <- ggplot(data = frog, aes(x = worms, y = fungus)) + geom_point()
myplot

  # add color and change data points
myplot <- ggplot(data = frog, aes(x = worms, y = fungus, color = sex)) + geom_point(size = 3)
myplot


## 3 geoms for summarization

# 1. Boxplot
ggplot(data = frog, aes(x = color, y = fungus)) + geom_boxplot()

# 2. Histogram; only need to specify 1 vector to be visualized in aes()
ggplot(data = frog, aes(x = worms)) + 
  geom_histogram(binwidth = 2, color = 'black', fill = 'white')

# 3. Facets; panels which plots of mapped variables are arranged according to a cateforical grouping variable
ggplot(data = frog, aes(x = worms, y = fungus)) + geom_point(size = 3) +
  facet_grid(.~sex) # sex is grouping variable
# specifying the group variable on the left 'facet_grid(sex~.)' arranges facets in one column
# specifying the group variable on the right 'facet_grid(.~sex)' arranges facets in one row
ggplot(data = frog, aes(x = worms, y = fungus)) + geom_point(size = 3) +
  facet_grid(sex~.)

# facet 2+ grouping variables
ggplot(data = frog, aes(x = worms, y = fungus)) + 
  geom_point(size = 3) +
  facet_grid(sex~color)

## Stats

# add linear fit line to a scatter plot
ggplot(data = frog, aes(x = worms, y = fungus)) + 
  geom_point(size = 3) +
  geom_smooth(method = lm)

# Theme and controlling figure appearance
# control the non-data elements (text, axes, legends)

ggplot(data = frog, aes(x = worms, y = fungus, color = virus)) + 
  geom_point(size = 3) +
  theme(legend.key = element_rect(fill = NA),
        legend.position = 'bottom', 
        axis.title = element_text(angle = 0, size = 18, face = 'bold'),
        legend.text = element_text(angle = 0, size = 12, face = 'bold'),
        panel.background = element_rect(fill = NA))

# save ggplot figure
ggsave('myplot.jpg', width = 8, height = 8, unit = 'cm', dpi = 300)
```


```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
