---
title: "Lab1"
author: "sbsambado"
date: "12/24/2020"
output: html_document
---

Part 1: We'll create a new GitHub repo and R project, wrangle and visualize data from spreadsheets in R Markdown, communicate between RStudio (locally) and GitHub (remotely) to keep our updates safe 

## Objectives

- Create a new repo on GitHub
- Clone the repo to create a new version-controlled R project 
- Create **data\** and **figures\** subfolder s
- Add noaa_fisheries.csv data to **data\** folder in working directory
- Create a new R Markdown document 
- Attach necessary packages (`tidyverse`, `here`, `janitor`, `kableExtra`)
- Use `here::here()` for simpler (and safer) file paths
- Data tidying and wrangling (`dplyr`, `tidyr`, etc.)
- HTML tables with `kable()` and `kableExtra`
- Data visualization (`ggplot2`)

## Resources

- [Project oriented workflows](https://www.tidyverse.org/blog/2017/12/workflow-vs-script/) by Jenny Bryan
- [R4DS](https://r4ds.had.co.nz/) by Garrett Grolemund and Hadley Wickham



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(here)
library(janitor)
library(kableExtra)

us_landings <- read_csv(here("data","noaa_fisheries.csv"))
```

```{r}
summary(us_landings)
View(us_landings)
names(us_landings)
head(us_landings)
tail(us_landings)
```

## Some data cleaning to get salmon landings

- Clean up column names (`clean_names()`)
- Convert everything to lower case with `mutate()` + (`str_to_lower()`)
- Remove dollar signs in value column (`mutate()` + `parse_number()`)
- Remove the "aggregate" grouping indicator in species (`mutate()` + a new function! `str_remove()`) %>% 
- Keep only observations that include "salmon" (`filter()`)
- Separate the grouped name ("salmon") from any additional refined information on species (`separate()`)

We'll break this up into two pieces:

1. A tidier version of the entire data frame
2. A subset that only contains salmon information

```{r}
# tidy whole dataframe
landings_tidy <- us_landings %>% 
  clean_names() %>% # make column headers snake_case
  mutate(
    state = str_to_lower(state),
    afs_name = str_to_lower(afs_name)
  ) %>% #converts character columns to lowercase
  mutate(dollars_num = parse_number(dollars_usd))

# just get salmon

salmon_landings <- landings_tidy %>% 
  mutate(afs_clean = str_remove(afs_name, pattern = "aggregate")) %>% 
  filter(str_detect(afs_clean, pattern = "salmon")) %>%  #detect matches for crab
  separate(afs_clean, into = c("group", "species"), sep = ",")

```

## Find grouped summary data

Find the annual total US landings and dollar value (summing across all states) for each type of salmon using `group_by()` + `summarize()`:
```{r}
salmon_summary <- salmon_landings %>% 
  group_by(year, species) %>% 
  summarize(
    tot_landings = sum(landings_pounds),
    tot_value = sum(dollars_num)
  )

salmon_landings_graph <- ggplot(salmon_summary, aes(x = year, y = tot_landings, group = species)) +
  geom_line(aes(color = species)) +
  theme_bw() +
  labs(x = "year", y = "US commercial salmon landings")

salmon_landings_graph

ggsave(plot = salmon_landings_graph, here("figures", "us_salmon_landings.png"),
       height = 5, width = 8)

```

## 2015 commercial fisheries value by state

Now, let's create a finalized table of the top 5 states (by total commercial fisheries value) for 2015 .

Remember that we already created a tidied data frame, **landings_tidy**. 

*Critical thinking for data wrangling workflow: Why does it make sense (especially now) that we broke our previous wrangling into two steps before getting our salmon subset?* 

```{r}
state_value <- landings_tidy %>% 
  filter(year %in% c(2015)) %>% 
  group_by(state) %>% 
  summarize(
    state_value = sum(dollars_num, na.rm = TRUE),
    state_landings = sum(landings_pounds, na.rm = TRUE)
  ) %>% 
  arrange(-state_value) %>% 
  head(5)


```


What are some ways we'd want to finalize it for a report?

- Change units (to millions of dollars and pounds)
- Update column names
- Capitalize the state names

## Making a nice HTML table

finalized data frame
```{r}

state_table <- state_value %>% 
  mutate('Fisheries value ($ millions)' = round(state_value / 1e6, 2),
         'Landings (million pounds)' = round(state_landings / 1e6, 1)) %>% 
  select(-state_value, - state_landings) %>% 
  rename(State = state) %>% 
  mutate(State = str_to_title(State))
```

Now, use `kable()` + `kableExtra` to nicely format it for HTML:

```{r}

kable(state_table) %>% 
  kable_styling(bootstrap_options = "striped",
                full_width = FALSE) %>% 
  add_header_above(c("", "2015 US commercial fisheries - top 5 states by value" = 2))

```


