---
title: "Lab2"
author: "sbsambado"
date: "12/24/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)

library(tidyverse)
library(janitor)
library(naniar)
library(VIM)
library(skimr)
library(ggfortify)

ca_pb <- read_csv("ca_pollution_burden.csv")
ca_dem <- read_csv("ca_census_demographics_2010.csv")
```


## B. Do some cleaning 

1. For the pollution burden data: 

- Clean up the column headers
- Exclude any column that is a calculated percentile (contains 'percentile', 'perc', or 'pctl')

```{r}
ca_pb_nopct <- ca_pb %>% 
  clean_names() %>% 
  select(-contains("pctl")) %>% 
  select(-contains("perc")) %>% 
  select(-latitude, - longitude)
```
2. For the demographic data:

- Clean up column names
```{r}
ca_dem_clean <- ca_dem %>% 
  clean_names()
```
3. PCA for pollution burden indicator variables

First, starting with ca_pb_nopct:

**Note**: The pollution burden and population characteristic variables are aggregates (averages) of existing variables in the data frame, so we won't include those. That means we'll include columns:

- From `ozone:solid_waste`, and
- Frome `asthma:housing_burden`

First, just selecting those:

```{r}
ca_pb_subset <- ca_pb_nopct %>% 
  select(ozone:solid_waste, asthma:housing_burden)
```

We'll use the `prcomp` function:
- ?prcomp

A little aside: the `naniar` package for exploring missingness!
See: https://naniar.njtierney.com/

Use `naniar::gg_miss_var()` to plot the number of missings by variable:

We can also explore NAs visually in other ways using the `VIM` package.

The `VIM::matrixplot()` function creates a heatmap, where values are shown on a continuous grayscale, and missings are in red:

Let's say our conclusion is that there are missings, but not many (compared to the actual scope of the data). We'll only keep our complete cases (census tracts without any missings).

Use `tidyr::drop_na()` with no variables specified to keep complete cases across all variables:

```{r, eval=FALSE}
# look at NA situation
gg_miss_var(ca_pb_subset)
matrixplot(ca_pb_subset)
# We can also sort by a variable of interest. Let's say we want to sort by poverty:
matrixplot(ca_pb_subset, sortby = "poverty")

# 
# summary(ca_pb_subset) # Max NAs in a variable: 242 (/8035)
# pc_pca <- prcomp(ca_pb_subset, scale = TRUE)

ca_pb_nona <- ca_pb_subset %>% 
  drop_na()

# check for NAs
summary(ca_pb_nona) 

# or use
skim(ca_pb_nona)
```

Cool. No NAs, checked out missingness, NOW let's try PCA again: 


```{r}

my_ca_pca <- prcomp(ca_pb_nona, scale = TRUE)

plot(my_ca_pca)
biplot(my_ca_pca)


my_biplot <- autoplot(my_ca_pca,
                      color = NA,
                      loading.label = TRUE,
                      loadings.label.size = 3,
                      loadings.label.color = "black",
                      loadings.label.repel = TRUE)+
  theme_minimal()

my_biplot
```

4. Join data by census tract (inner join)
```{r}
ca_df <- ca_dem_clean %>% 
  inner_join(ca_pb_nopct, by = c("census_tract_number" = "census_tract"))

# get complete cases

ca_df_nona <- ca_df %>% 
  drop_na()
```

5. Make a new subset for PCA, that includes % white and elderly, and some interesting pollution burden & health indicators:

Like (you can choose a different set): 

- white_percent
- elderly_65_percent
- pm2_5
- pesticides
- traffic
- asthma
- cardiovascular_disease
- poverty

```{r}
# make subset

my_sub <- ca_df_nona %>% 
  select(white_percent, elderly_65_percent, pm2_5, pesticides, traffic, asthma, cardiovascular_disease, poverty)

# run the PCA

my_dem_pca <- prcomp(my_sub, scale = TRUE)

biplot(my_dem_pca)

# Proportion of variance (& cumulative variance) explained by each PC
summary(my_dem_pca)
# Rotations (linear combinations for each PC):
my_dem_pca


# make a sweet biplot

my_dem_biplot <- autoplot(my_dem_pca,
                          color = NA, 
                          loadings.label = TRUE,
                          loadings.label.size = 3,
                          loadings.label.color = "black",
                          loadings.label.repel = TRUE)+
  theme_minimal()+
  scale_y_continuous(limits = c(-0.05, 0.05))

my_dem_biplot
```

