---
title: "Lab5_Timeseries"
author: "sbsambado"
date: "1/10/2021"
output: html_document
---

Lab Week 5 Objectives:

- Intro to time series data class (ts)
- Graphing ts data
- Decomposing ts data
- Exploring autocorrelation
- Holt Winters and ARIMA preview


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

TimeSeriesData <- ts(DataVector,
frequency = ?,
start c(Year, IntervalStart))

frequency = 
12: Monthly Data
4: Quarterly Data
7: Weekly Data

or deltat = ?

1/52: Weekly
1/365: Daily

```{r}
library(tidyverse)
library(tseries)
library(forecast)
```

```{r}

energy <- read_csv("energy.csv")

# convert to ts data

res_ts <- ts(energy$res_total, frequency = 12,
              start = c(1973, 1))

plot(res_ts)
```
For each, we should ask ourselves:
- Is there a trend?
- Do data look additive or multiplicative? 
- Is there seasonality? 
- Are there notable outliers? 

d. Decompose to start exploring the data further
```{r}
res_dc <- decompose(rest_ts)
plot(res_dc)

# Changes within each month over years recorded:
monthplot(res_ts)

ggseasonplot(res_ts)+
  theme_bw()


```


f. Simple moving average to smooth (changing the averaging window)
```{r}
# Have them see what happens when they change the moving window...

sma_res <- ma(rest_ts, order = 5)
plot(sma_res)

# Just view the moving average on its own: 
plot(sma_res)
# Or in combo with the original data
plot(res_ts)
lines(sma_res, col = "red")
```

g. Exploring autocorrelation (ACF) - two ways
```{r}

res_acf <- acf(res_ts)

ggtsdisplay(res_ts)

#Not surprising: strong seasonality is dominant. There appears to be some trend. It looks relatively additive. Can we test for stationarity? 

```

h. Augmented Dickey-Fuller test for stationarity

Hypothesis test: null is that the data are NOT stationary. If p < 0.05, we reject the null hypothesis and retain the alternative hypothesis that the data ARE stationary.
```{r}
adf_res <- adf.test(res_ts)# Yes, stationary
adf_res  # p-value = 0.01
```

i. Holt Winters exponential smoothing
```{r}
# Exponential smoothing: no normality assumption (unbiased)
# Perform Holt Winters

res_hw <- HoltWinters(res_ts)# See smoothing parameters with res_hw
plot(res_hw)

# Then forecast
res_forecast <- forecast(res_hw, h = 60)
plot(res_forecast)

#Then check the residuals:
hist(res_forecast$residuals) # looks normal distributed
```

j. Autoregressive integrated moving average (ARIMA) using auto.arima for p, d, q

- Use auto.arima to estimate pdq
```{r}
res_pdq <- auto.arima(res_ts) # [1,0,2][0,1,1]
res_pdq
```

- Fit the ARIMA model
```{r}
res_arima <- arima(res_ts, order = c(1,0,2),
                   seasonal = list(order = c(0,1,1)))
```

- Evaluate residuals (look good)
```{r}
par(mfrow = c(1,2))
hist(res_arima$residuals)
qqnorm(res_arima$residuals)
```

- Look at the forecasting...
```{r}

forecast_res <- forecast(res_arima, h = 72)
plot(forecast_res)
```

- And view this in ggplot (if time)?
```{r}
res_df <- data.frame(forecast_res)
month_seq <- seq(1,72)

res_df_2 <- data.frame(month_seq, res_df)

ggplot(res_df_2, aes(x = month_seq, y = Point.Forecast))+
  geom_line()+
  geom_ribbon(aes(ymin = Lo.95, ymax = Hi.95, alpha = .2))+
  theme_minimal()
```

