---
title: "Lab3"
author: "sbsambado"
date: "12/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


```

If knitting to PDF, DEFINITELY set message = FALSE in the chunk where you load the tidyverse, because there is a character in the message that LaTex doesn't understand and you'll get an error. Alternatively, use suppressMessages(library(tidyverse))

```{r, echo = FALSE, message = FALSE}

library(tidyverse)
library(ggpubr)

flint <- read_csv("flint.csv")

```

##A. Probabilities and confidence intervals for the normal distribution

If we have satisfied assumptions to use the Z-distribution, we can use built-in functions in R to calculate probabilities associated with it. The ones we'll use today are *pnorm* and *qnorm*. 

*pnorm* is used to find probabilities (areas under the curve)...the default is finding the area to the LEFT of the value you're interested in (lower tail)

*qnorm* is used to find the value of the random variable associated with a given percentile (probability)

Let's say we're told the following: You are told that the population distribution of cat weights (pounds) is normally distributed with a mean of 9.2 pounds and a standard deviation of 1.4 pounds. 

We ask:

1. Why would it be appropriate to use the Z-distribution to calculate probabilities? 

2. What is the probability of randomly selecting a cat from the population that:

- weighs less than 8 pounds?
- weighs more than 10 pounds? 
- weighs between 8 and 11 pounds?

When in doubt, always draw a picture!

Heads up here: if knitting to a PDF, text in code doesn't automatically wrap (and there's actually not a really easy way to do it).


Probabilities with Z-distribution
```{r}


cat_mean <- 9.2
cat_sd <- 1.4

less_8 <- pnorm(8, 
                mean = cat_mean,
                sd = cat_sd)

more_10 <- pnorm(10,
                 mean = cat_mean,
                 sd = cat_sd, 
                 lower.tail = FALSE)

between_9_10 <- pnorm(11, mean = cat_mean, sd = cat_sd) -
  pnorm(8, mean = cat_mean, sd = cat_sd)
```

3. At what weight would a cat be at the 90th percentile in weight for the population? At the 30th percentile?

```{r}

perc_90 <- qnorm(0.9,
                 mean = cat_mean,
                 sd = cat_sd)

perc_30 <- qnorm(.3, 
                 mean = cat_mean,
                 sd = cat_sd)
```


#B. The t-distribution for confidence intervals and hypothesis test intro (one-sample t-test)

When we use pnorm/dnorm we're evaluating probabilities using the Z-distribution. What if we don't know the population sd? If our sample is normally distributed (even at n < 30), or if we have a larger sample (n > 30, due to CLT), we can still use the t-distribution to evaluate probabilities. 

We'll do that in the context of a one-sample hypothesis test to test a claim as an example, using the *t.test()* function. 


```{r}


ggplot(iris, aes(sample = Petal.Length)) +
  geom_qq(aes(color = Species)) +
  facet_wrap(~ Species, scale = "free")


ggplot(iris, aes(x = Petal.Length)) +
  geom_histogram(aes(fill = Species), bins = 8) +
  facet_wrap(~Species, scale = "free")

```
We read a claim that the mean petal length of setosa irises is 1.5cm. We question that claim, and luckily have Edgar Anderson's iris data to use as a sample. Based on Anderson's measurements, do we have enough evidence to reject the claim? 


a. First, create a subset that only contains information for setosa petal lengths:

```{r}

setosa_petals <- iris %>% 
  filter(Species == "setosa") %>% 
  select(Petal.Length) %>% 
  rename(length = Petal.Length)
```


b. Next, do some exploring! This is always necessary.
```{r}

hist(setosa_petals$length)
qqnorm(setosa_petals$length)


ggplot(setosa_petals, aes(sample = length)) +
  geom_qq()

mean_pl <- mean(setosa_petals$length)
sd_pl <- sd(setosa_petals$length)
```

c. Find the 95% confidence interval for setosa iris petal length. What does that mean?

```{r}


setosa_ci <- t.test(setosa_petals$length)
setosa_ci

# 95 percent confidence interval:
#  1.412645 1.511355
# "Mean setosa petal length (n = 50) is 1.46 cm, with a 95% confidence interval of [1.41, 1.51] cm." 

# What if we wanted to find an 80% confidence interval?

setosa_ci_80 <- t.test(setosa_petals$length, conf.level = 0.8)
setosa_ci_80
```

d. Now, we'll test the claim using a two-sided, one-sample t-test.

Null hypothesis: Mean petal length for setosa irises is 1.5cm.
Alternative hypothesis: Mean petal length is NOT 1.5cm (two-sided).

```{r}

tpetal <- t.test(setosa_petals$length, mu = 1.5) # p = 0.12 (p > 0.05)

# There is insufficient evidence to reject the claim that mean setosa iris petal length is 1.5 cm (t(49) = -1.55, p = 0.13)
```

e. What if we read a claim that the mean petal length is 1.6 cm, but we think that the true mean is LOWER than 1.6 cm?

```{r}

tpetal_lower <- t.test(setosa_petals$length, mu = 1.6, alternative = "less") # p 4.5e-07 ( p < 0.05)

# If the true mean is 1.6cm, there is only a probability of 4.5x10^-7^ that a sample (n = 50) drawn from that population would have a mean of OUR sample mean (1.462cm) or less by random chance...it's more likely that the claim is wrong, and that the true mean is actually lower. 

# Based on our sample (mean = 1.46, n = 50), we conclude that the mean petal length for setosa irises is lower than the claim of 1.6 cm (t(49) = -5.6, p < 0.001).


```


#C. Two-sample t-tests (Flint water)

Describe dataset
```{r}


pb <- flint %>% 
  select(pb_1, pb_3)

summary(pb)

pb1hist <- ggplot(pb, aes(x = pb_1)) +
  geom_histogram()

pb1qq <- ggplot(pb, aes(sample = pb_1)) +
  geom_qq()

pb3hist <- ggplot(pb, aes(x = pb_3)) +
  geom_histogram()

pb3qq <- ggplot(pb, aes(sample = pb_3)) +
  geom_qq()

ggarrange(pb1hist, pb1qq, pb3hist, pb3qq)
```

# These are really non-normally distributed...why would we still be able to use the t-distribution to compare *means* (assuming that we think *means* is an interesting thing to compare...)

b. T-test
```{r}

flint_t <- t.test(pb$pb_1, pb$pb_3, alternative = "greater",
                  paired = TRUE)
flint_t  # p = 3.946e-10 (p < 0.05)

# (t(270) = 6.3748 , p < 0.001)

#Are lead concentrations significantly reduced post-flushing? YES...by a LOT. Post-flushing Pb concentrations are significantly lower than pre-flushing concentrations.



```

