---
title: "Lab7"
author: "sbsambado"
date: "12/11/2020"
output: html_document
---

random fact

read_csv reads in tibbles

+ load faster
+ don't change input types
+ allow you to have columns as lists
+ allow non-standard variable names (variables can start with number and contain spaces)
+ never create row names


read.csv reads in dataframe

**you should load tibble instead of data frame** 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(magrittr)

gc_df <- read_csv("gc_df.csv")
kermit <- read_csv("kermit.csv")


```

```{r}

get_guns <- gc_df %>% 
  filter(gender != "NA", make_get_guns != "NA") %>% 
  count(gender, make_get_guns) %>% 
  spread(make_get_guns, n) %>% 
  select(-gender)

rownames(get_guns) <- c("Female", "Male")

# Tell it what variable you want to spread out into multiple columns (KEY), and what the values should be associated with it (Freq)...this is called a 'key-value pair'
# Now this looks like a contingency table. 

```

# What we're really asking: 
Do PROPORTIONS of respondents who choose "make it more difficult","make it less difficult", or "same" to get guns differ between males and females?

# First, let's look at the actual proportions:

# Could also do this using the tidyverse, using group_by + mutate, but we'll use prop.table (which really wants thing in the class of 'matrix'):

```{r}

guns_prop <- round(prop.table(as.matrix(get_guns),1),3)

# Perform chi-square test for independence (on the COUNT data - NOT on the proportions!)
guns_x2 <- chisq.test(get_guns)
guns_x2 # p < 2.2e-16

guns_x2$stdres # If beyond 2, pay attention for significance...
# There IS a significant association between response and gender! Look at 'Values' section to see what things you can call from the output.

"Recommended government action on gun purchasing differed significantly between male and female respondents ($\chi^2$(`r guns_x2$parameter`) = `r guns_x2$statistic`, *p* = `guns_x2$p.value`, $\alpha$ = 0.05. Most notably, a much higher proportion of female respondents (57.4%) recommend making it more difficult to purchase guns, compared to 40.4% for male respondents."

```


Rank based tests

```{r}

kermit_hist <- ggplot(kermit, aes(x = Score.pre)) +
  geom_histogram(bins = 4) +
  facet_wrap(~ Gender)

kermit_hist

# perform Mann-Whitney U to compare pre-presentation kermit rankings between M / F respondents:

mwu_kermit <- kermit %>% 
  wilcox.test(Score.pre ~ Gender, data = .)
mwu_kermit # p = .4001 (p > 0.05)

# There is no significant difference in kermit ranks for Male and Female respondents. 
#Ranks comparison by Mann-Whitney U reveals no significant difference in pre-presentation Kermit ranks for male and female respondents (W = `r mwu_kermit$statistic`, *p* = `r mwu_kermit$p.value`, $\alpha = 0.05$)

```

Wilcoxon Signed Rank (paired comparison of ranked data). 

Question: "Is there a significant difference in pre- and post-presentation kermit rankings?"

```{r}
wsr_kermit <- wilcox.test(kermit$Score.pre, kermit$Score.post, paired = TRUE)
wsr_kermit # p = 0.3143 (p > 0.05)

# There is NO significant difference in rankings before and after the presentation. 

```


