---
title: "Chapter 13"
author: "Stephen R. Proulx"
date: "5/19/2020"
output: html_document
---

```{r setup}

knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(rstan)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())


library(rethinking)

```


```{r}
data(reedfrogs)
d <- as_tibble(reedfrogs)
d$tank <- 1:nrow(d)
```

ulam model
```{r}
dat <- list(
  S=d$surv,
  N= d$density,
  tank = d$tank)

m13.1 <- ulam(
  alist(
    S ~ dbinom( N, p),
    logit(p) <- a[tank],
    a[tank] ~ dnorm(0,1.5)
  ),
  data=dat, chains=4, log_lik = TRUE
  )
```

```{r}
precis(m13.1 , depth =2)
```



## Use brms to fit the models-
```{r}
detach(package:rethinking, unload = T)
library(brms)
```


A brms model for the tanks. Here we fit them all in a single model. We'll compare it to a fit of a subset, and see that it comes out the same because they are all effectively being fit alone.

Note here using the formula  0 + ... gets  brms to not have an intercept tank
```{r}

d <- mutate(d, tank = as.factor(tank))


tadpole_tanks <- brm(surv | trials(density) ~ 0 + tank ,
                     family = binomial(logit),
                     data=d,
                     prior = c(prior(normal(0,1.5), class = b)) )
```



```{r}
print(tadpole_tanks)
```

```{r}
tadpole_marginal <- marginal_effects(tadpole_tanks) 


plot(tadpole_marginal)[[1]]+
  ggplot2::scale_y_continuous( limits=c(0,1),breaks=seq(0,1,0.1), labels=seq(0,1,0.1))
```
now redo the analysis with just the first 5 tanks, basically we get the same result
```{r}

d$tank_num <- 1:nrow(d)
 

tadpole_tanks_sub <- update(tadpole_tanks,
                            newdata = filter(d,tank_num<=5))

tadpole_marginal_sub = marginal_effects(tadpole_tanks_sub)



plot(tadpole_marginal_sub)[[1]]+
  ggplot2::scale_y_continuous( limits=c(0,1),breaks=seq(0,1,0.1), labels=seq(0,1,0.1))
```

## Simulating the data and showing advantage of partial pooling
Side note: using ulam from here on out, I updated some packages and broke part of tidyverse and brms...

Define the parameters to use in this simulation
```{r}
## R code 13.7
a_bar <- 1.5
sigma <- 1.5
nponds <- 60
Ni <- as.integer( rep( c(5,10,25,35) , each=15 ) )
```

Choose the "actual" binomial parameter for each pond
```{r}
## R code 13.8
set.seed(5005)
a_pond <- rnorm( nponds , mean=a_bar , sd=sigma )
 
```


```{r}
## R code 13.9
dsim <- tibble( pond=1:nponds , Ni=Ni , true_a=a_pond )  
 
## R code 13.11
dsim$Si <- rbinom( nponds , prob=logistic(dsim$true_a) , size=dsim$Ni )
```


Put in the no pooling estimate, which is the maximum likelihood estimate and is just the actual percentage that survived.
```{r}
## R code 13.12
dsim$p_nopool <- dsim$Si / dsim$Ni
```


Run the mutli-level model.
```{r}
## R code 13.13
dat <- list( Si=dsim$Si , Ni=dsim$Ni , pond=dsim$pond )
m13.3 <- ulam(
    alist(
        Si ~ dbinom( Ni , p ),
        logit(p) <- a_pond[pond],
        a_pond[pond] ~ dnorm( a_bar , sigma ),
        a_bar ~ dnorm( 0 , 1.5 ),
        sigma ~ dexp( 1 )
    ), data=dat , chains=4 )
```


```{r}
## R code 13.14

plot(m13.3 , depth=2)

precis( m13.3 , depth=2 )

```

Extract the samples from the model, convert them to the probability scale, and caclulate the error from true
```{r}
## R code 13.15
post <- extract.samples( m13.3 ) 

dsim$p_partpool <- apply( inv_logit(post$a_pond) , 2 , mean )

## R code 13.16
dsim$p_true <- inv_logit( dsim$true_a )
```


Take a minute to look at the error.
```{r}
view(dsim)
```


add the error measurement and plot
```{r}
## R code 13.17
dsim <- dsim %>%
  mutate(nopool_error = abs(p_nopool-p_true), partpool_error = abs(p_partpool-p_true))


dfline <- 
  dsim %>%
  select(Ni, nopool_error:partpool_error) %>%
  gather(key, value, -Ni) %>%
  group_by(key, Ni) %>%
  summarise(mean_error = mean(value)) %>%
  mutate(x    = c( 1, 16, 31, 46),
         xend = c(15, 30, 45, 60))


 ggplot(dsim,aes(x = pond)) +
  geom_vline(xintercept = c(15.5, 30.5, 45.4), 
             color = "white", size = 2/3) +
  geom_point(aes(y = nopool_error), color = "orange2") +
  geom_point(aes(y = partpool_error), shape = 1) +
  geom_segment(data = dfline, 
               aes(x = x, xend = xend, 
                   y = mean_error, yend = mean_error),
               color = rep(c("orange2", "black"), each = 4),
               linetype = rep(1:2, each = 4)) +
  scale_x_continuous(breaks = c(1, 10, 20, 30, 40, 50, 60)) +
  annotate("text", x = c(15 - 7.5, 30 - 7.5, 45 - 7.5, 60 - 7.5), y = .45, 
           label = c("tiny (5)", "small (10)", "medium (25)", "large (35)")) +
  labs(y        = "absolute error",
       x = "pond number")+
  theme(panel.grid    = element_blank(),
        plot.subtitle = element_text(size = 10))
```

Re-run for a new simulation without re-compiling
```{r}
## R code 13.20
a <- 1.5
sigma <- 1.5
nponds <- 60
Ni <- as.integer( rep( c(5,10,25,35) , each=15 ) )
a_pond <- rnorm( nponds , mean=a , sd=sigma )
dsim <- data.frame( pond=1:nponds , Ni=Ni , true_a=a_pond )
dsim$Si <- rbinom( nponds,prob=inv_logit( dsim$true_a ),size=dsim$Ni )
dsim$p_nopool <- dsim$Si / dsim$Ni
newdat <- list(Si=dsim$Si,Ni=dsim$Ni,pond=1:nponds)
m13.3new <- stan( fit=m13.3@stanfit , data=newdat , chains=4 )

post <- extract.samples( m13.3new )
dsim$p_partpool <- apply( inv_logit(post$a_pond) , 2 , mean )
dsim$p_true <- inv_logit( dsim$true_a )
dsim <- dsim %>%
  mutate(nopool_error = abs(p_nopool-p_true), partpool_error = abs(p_partpool-p_true))


dfline <- 
  dsim %>%
  select(Ni, nopool_error:partpool_error) %>%
  gather(key, value, -Ni) %>%
  group_by(key, Ni) %>%
  summarise(mean_error = mean(value)) %>%
  mutate(x    = c( 1, 16, 31, 46),
         xend = c(15, 30, 45, 60))


 ggplot(dsim,aes(x = pond)) +
  geom_vline(xintercept = c(15.5, 30.5, 45.4), 
             color = "white", size = 2/3) +
  geom_point(aes(y = nopool_error), color = "orange2") +
  geom_point(aes(y = partpool_error), shape = 1) +
  geom_segment(data = dfline, 
               aes(x = x, xend = xend, 
                   y = mean_error, yend = mean_error),
               color = rep(c("orange2", "black"), each = 4),
               linetype = rep(1:2, each = 4)) +
  scale_x_continuous(breaks = c(1, 10, 20, 30, 40, 50, 60)) +
  annotate("text", x = c(15 - 7.5, 30 - 7.5, 45 - 7.5, 60 - 7.5), y = .45, 
           label = c("tiny (5)", "small (10)", "medium (25)", "large (35)")) +
  labs(y        = "absolute error",
       x = "pond number")+
  theme(panel.grid    = element_blank(),
        plot.subtitle = element_text(size = 10))
```



