---
title: "Chapter 7/8 Interactions"
author: "Stephen R. Proulx"
date: "5/4/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(rstan)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
library(brms)

library(rethinking)
```


```{r load_process_data}
data(rugged)
d <- as_tibble(rugged)


# make log version of outcome and  extract countries with GDP data
d <- d %>% 
  mutate(log_gdp = log(rgdppc_2000))%>%
  drop_na(rgdppc_2000) %>% 
  mutate(log_gdp_std=log_gdp/mean(log_gdp), rugged_std = rugged/max(rugged))
```

Fit separate models for the within Africa and other countries
```{r}
b7.1 <-
  brm(data = filter(d ,cont_africa == 1), # in africa
      family = gaussian,
      log_gdp ~ 1 + rugged, # log of lgd is intercept + how rugged the country is
      prior = c(prior(normal(1,1), class = Intercept),
                prior(normal(0, 1), class = b),
                prior(cauchy(0,10), class = sigma)), # how positive
      iter = 2000, warmup = 1000, chains = 4, cores = 4,# parameters of sampling run
      seed = 2312312) # repeatability

b7.2 <- # out of africa
  update(b7.1, # take stan object associated with it (binary program)
         newdata = filter(d ,cont_africa == 0)) # replaces with new data
 
```


```{r}
print(b7.1)
print(b7.2)
```

```{r}
nd <- 
  tibble(rugged = seq(from = 0, to = 6.3, length.out = 30))

f_b7.1 <- # fitted fxn is convenience fxn, does the work for ya
  fitted(b7.1, newdata = nd) %>% # each point calculate mean in likelihood fxn for ruggedness value you're testing out, gives output that is a table for each sample value of mean for sampling distribution, take that mean
  as_tibble() %>%
  mutate(cont_africa = 1)%>%
  bind_cols(nd) 

f_b7.2 <-
  fitted(b7.2, newdata = nd) %>%
  as_tibble() %>%
  mutate(cont_africa =0)%>%
  bind_cols(nd)

#put them back into a single data-frame for plotting purposes
f <-
  bind_rows(f_b7.1, f_b7.2) 

 

ggplot(d,aes(x = rugged)) +
  geom_smooth(data = f,
              aes(y = Estimate, ymin = Q2.5, ymax = Q97.5,
                  fill = as.factor(cont_africa), color = as.factor(cont_africa)),
              stat = "identity", 
              alpha = 1/4, size = 1/2)+
  geom_point(aes(y = log_gdp, color = as.factor(cont_africa)),
             size = 2/3) +
  scale_x_continuous("Terrain Ruggedness Index", expand = c(0, 0)) +
  ylab("log GDP from year 2000") +
  theme(text = element_text(family = "Times"))  

```
Run the interaction model
- put all 3 group models together
```{r}
b7.3 <-
  brm(data = d,  # whole data set
      family = gaussian,
      log_gdp ~ 1 + rugged , # same intercept
      prior = c(prior(normal(1,1), class = Intercept),
                prior(normal(0, 1), class = b),
                prior(cauchy(0,10), class = sigma)),
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      seed = 2312312)


b7.4 <-
  brm(data = d, # in africa
      family = gaussian,
      log_gdp ~ 1 + rugged + cont_africa, #same slope for ruggedness, different intercept
      prior = c(prior(normal(1,1), class = Intercept),
                prior(normal(0, 1), class = b),
                prior(cauchy(0,10), class = sigma)),
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      seed = 2312312)

b7.5 <-
  brm(data = d,  # out of africa
      family = gaussian,
      log_gdp ~ 1 + rugged * cont_africa, # multiplying two factors to mean that both of them have an effective. Global slope, global intercept, difference in slope if you're in africa vs. outside of it
      prior = c(prior(normal(1,1), class = Intercept),
                prior(normal(0, 1), class = b),
                prior(cauchy(0,10), class = sigma)),
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      seed = 2312312)


```

```{r}
print(b7.3)
print(b7.4)
print(b7.5)

```
^ standard linear model focuses on reference case (way 1)
^ brms plots is a more direct description of the linked model (way 2)

Add information criterion to the model fits and compare!
- output for both ways are the same in the way you plot them 
```{r}
b7.3<-add_criterion(b7.3,c("loo","waic"))
b7.4<-add_criterion(b7.4,c("loo","waic"))
b7.5<-add_criterion(b7.5,c("loo","waic"))


l<-loo_compare(b7.3,b7.4,b7.5,
            criterion = "waic")

print(l, simplify = F)


```


Compare the interaction model with the "run them separately model"
```{r}
nd1 <- 
  tibble(rugged = seq(from = 0, to = 6.3, length.out = 30),cont_africa=0)
nd2 <- 
  tibble(rugged = seq(from = 0, to = 6.3, length.out = 30),cont_africa=1)


f_b7.5_A <-
  fitted(b7.5, newdata = nd1) %>%
  as_tibble() %>%  bind_cols(nd1)

f_b7.5_N <-
  fitted(b7.5, newdata = nd2) %>%
  as_tibble() %>%
  bind_cols(nd2)

#put them back into a single data-frame for plotting purposes
f_7.5 <-
  bind_rows(f_b7.5_A, f_b7.5_N) 




ggplot(d,aes(x = rugged)) +
  geom_smooth(data = f_7.5,
              aes(y = Estimate, ymin = Q2.5, ymax = Q97.5,
                  fill = as.factor(cont_africa), color = as.factor(cont_africa)),
              stat = "identity", 
              alpha = 1/4, size = 1/2)+
  geom_point(aes(y = log_gdp, color = as.factor(cont_africa)),
             size = 2/3) +
  scale_x_continuous("Terrain Ruggedness Index", expand = c(0, 0)) +
  ylab("log GDP from year 2000") +
  theme(text = element_text(family = "Times"))  


ggplot(d,aes(x = rugged)) +
  geom_smooth(data = f_7.5,
              aes(y = Estimate, ymin = Q2.5, ymax = Q97.5,
                 fill = as.factor(cont_africa), color = as.factor(cont_africa+2)),
              stat = "identity") +
    geom_smooth(data = f, # just predictions and intervals for the 2 ways to do this
              aes(y = Estimate, ymin = Q2.5, ymax = Q97.5,
                  fill = as.factor(cont_africa), color = as.factor(cont_africa)),
              stat = "identity")+ scale_x_continuous("Terrain Ruggedness Index", expand = c(0, 0)) +
  ylab("log GDP from year 2000") +
  theme(text = element_text(family = "Times"))  




```




##continuous interaction data
Tulip data

```{r}
data(tulips)
d <- tulips

d <-
  d %>%
  mutate(shade_c = shade - mean(shade),
         water_c = water - mean(water))

```

```{r}
b7.8 <-
  brm(data = d, family = gaussian,
      blooms ~ 1 + water_c + shade_c,
      prior = c(prior(normal(130, 100), class = Intercept),
                prior(normal(0, 100), class = b),
                prior(cauchy(0, 10), class = sigma)),
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      control = list(adapt_delta = 0.9),
      seed = 7)

b7.9 <- 
  update(b7.8, 
         formula = blooms ~ 1 + water_c + shade_c + water_c:shade_c)
```

```{r}
print(b7.8)
print(b7.9)

b7.8<-add_criterion(b7.8,c("loo","waic"))
b7.9<-add_criterion(b7.9,c("loo","waic"))

loo_compare(b7.8,b7.9)

```


```{r}
# loop over values of `water_c` and plot predictions
shade_seq <- -1:1

for(w in -1:1){
  # define the subset of the original data
  dt <- d[d$water_c == w, ]
  # defining the data to be used for producing the model fit predictions
  nd <- tibble(water_c = w, shade_c = shade_seq)

    # Sample from the psoterior to produce fits for the mean values
  f <- 
    fitted(b7.9, newdata = nd) %>%
    as_tibble() %>%
    bind_cols(nd)
  
  # specify our custom plot
  fig <- 
    ggplot() +
    geom_smooth(data = f,
                aes(x = shade_c, y = Estimate, ymin = Q2.5, ymax = Q97.5),
                stat = "identity", 
                fill = "#CC79A7", color = "#CC79A7", alpha = 1/5, size = 1/2) +
    geom_point(data = dt, 
               aes(x = shade_c, y = blooms),
               shape = 1, color = "#CC79A7") +
    coord_cartesian(xlim = range(d$shade_c), 
                    ylim = range(d$blooms)) +
    scale_x_continuous("Shade (centered)", breaks = c(-1, 0, 1)) +
    labs("Blooms", 
         title = paste("Water (centered) =", w)) +
    theme(text = element_text(family = "Times"))
  
  # plot that joint
  plot(fig)
}
```


