---
title: "Applied Statistics Qualifying Exam"
author: "Samantha Sambado (Perm # 9917055)"
date: "Spring 2022"
output:
  pdf_document: default
  html_document: default
---



## Botts' dots impact on car crashes - simple solution or part of a complex chain of circumstances?

## Executive Summary

The postwar boom in the 1950s led to rapid increase of cars and collisions which in turn has spurred the development of Botts' dots as a simple solution to a complex problem. Botts' dots are now being phased out of some states, particularly California, where citizens, city planners, and engineers are wondering the impact of their removal on car collisions. In this report I evaluated the effectiveness of Botts' dots on count of car crashes across 1,000 road segments during the 1980s. With a Poisson generalized linear model I find that roads that do not have Botts' dots tend to have more car crashes, however other variables such as average traffic volume, driver's speed, and amount of drive ways were also signficant in my analyses. I recommend additional analyses that incorporate a driver's age and gender as well as weather conditions the day of crashes be investigated before people worry about the future removal of Botts' dots or incorporating other safety infrastructure.

## Introduction

Traffic accidents are one of the leading contributors to unintentional injury deaths in the United States and resulted in 36,096 deaths in 2019 alone (**Olaisen 2019**). Although there has been an overall decline in motor vehicle traffic death rates from 1999 through 2019, traffic death rates have increased in recent years (**Spencer 2021**). Technologies to reduce traffic accidents and ultimately reduce traffic deaths have gained a lot of interest of city planners and commuters with one notable example being Bott's dots.

Botts' dots (also known as turtles in Oregon or buttons in Texas) are round non-reflective raised pavement markers named after Elbert Dysart Botts, a California Department of Transportation (Caltran) engineer who oversaw the research to develop the markers (**Wikipedia**). The purpose of the Botts' dots are to mark lanes on highways and large roads to provide tactile and auditory feedback to drivers when moving across designated travel lanes. In California there are more than 25 million Bott's dots in use but they have begun to fall out of favor by Caltrans due to safety of workers and desire to make roadways more compatible with self-driving cars. 

This report will evaluate the effectiveness of Botts' dots in reducing traffic accident based on previously collected data about the number of traffic accidnets and road information on 1,000 road segments in California during 1980's when Botts' dots were installed. The dataset, CalCrash.csv, is a retrospective study that is comprised of 10,000 individual rows and 14 columns and has recorded 15,560 car crashes. To address the specific question of Botts' dots effectiveness in reducing traffic accidents the outcome of interest is `crash` which is a numeric count of the total number of crashes on a particular road segment at a particular time point. The main effect that will be investigated in relation to `crash` will be `botts` which is a binomial factor of 1 (yes) or 0 (no) as an indicator of whether Botts' dots were installed. With some a priori knowledge about traffic accidents, other effects that may influence `crash` will be investigated throughout this report such as `adt` (annual average daily traffic volume in the unit vehicles per day), `speed` (posted speed limit that is coded as a factor for two levels < 55 or > 55), `intersections` (factor with two levels no intersection or 1+ intersections), and `location` (road site index). There are other variables in this dataset that may not have a direct, fixed effect on `crash` but may interact with other fixed effects such as `length` (roadway segment length in miles), `width` (pavement width as a factor with three levels < 20, 20 - 24, > 24), `shoulder` (average shoulder width as a factor with three levels < 3, 3 - 6, > 6), `driveways` (number of driveways as a factor with two levels 1 - 10 or > 11), `curvature` (average degree of curvature), and `year` (calendar year in which the road site information was recorded).

For the M.S. pass I will assume that the road conditions did not change during 1980's or investigate if a trend exists over time. 


```{r setup, include=FALSE}

# Upload packages & data

knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)


library(dplyr)
library(readr)
library(ggplot2)
library(tidyverse)
library(psych)
library(car)
library(MASS)
library(jtools)
library(effects)
library(kableExtra)
library(ggpubr)
library(lme4)
library(boot)
library(patchwork)


# upload 
crash <- read.csv("qual_exam/CalCrash.csv")
str(crash) # 10,000 rows x 14 columns

# check data types

crash$botts <- as.factor(crash$botts)
crash$width <- as.factor(crash$width)
crash$speed <- as.factor(crash$speed)
crash$shoulder <- as.factor(crash$shoulder)
crash$driveways <- factor(crash$driveways,
                             levels = c("None", "> 0 and <= 10", "> 10"))
crash$intersections <- as.factor(crash$intersections)
crash$curves <- as.factor(crash$curves)
crash$year <- as.factor(crash$year)
crash$location <- as.factor(crash$location)

# check they changed
str(crash)

# check levels of variables
unique(crash$crash) # 0  1  2  7  5  4  3 11 12  6 14  9  8 13 10 18 15 16 17
unique(crash$year) # 10 years from 1980 - 1989

# reorder factors


```

## Exploratory Data Analysis

To explore the data before building models I *(1)* looked at the normality distribution of `crash`, *(2)* evaluated the relationship between `crash` and my fixed variables of interest, and *(3)* explored those relationships with the additional variable of `bott`.

To determine the normality of `crash` I made a histogram that had a positive, right skew which is to be expected for count data (**Fig. 1**). I then tried two transformations of `crash` such as square root and log transformation, however those transformations did not improve the normality of `crash` (**Appendix Part A**). This was also seen with a qqPlot where a step pattern was present and most points did not fall along the 1:1 normality line (**Appendix Part A**). However, because there are 10,000 data points within this dataset, The Central Limit Theorem can be applied since normality may be assumed when the dataset is a sufficiently large random samples from the population.

```{r fig.cap= "Normality check of car crash counts", fig.width= 4, fig.height= 2}
### look at normality

## histogram
# untransformed
ggplot(crash, aes(x = crash)) +
  geom_histogram()+
  theme_bw() +
  theme(axis.title = element_text(face = "bold")) +
  labs(x = "Counts of Car Crashes", y = "Frequencies")

```



To evaluate the relationship between `crash` and other individual co-variates, I first made a `pairs.panel()` plot of matrices with bivariate scatter plots that also calculates the Pearson correlation to help guide which variables should be focused on since the dataset has many variables (**Appendix Part B**). From this plot I determined potential relationships between `crash` and `adt` ($\rho$ = 0.54) (**Fig. 2**), then potentially `width` ($\rho$ = 0.16), `shoulder` ($\rho$ = 0.16). With bivariate scatter plots for crash as the response variable, `adt` has a positive linear trend as well as positive trend for length over 0.6 miles (**Appendix Part B**). Looking at `crash` as a response variable to factors `driveways` (level > 10), and `shoulder` (level > 6) had varying means of crash counts based on individual levels (**Appendix Part B**). 

```{r, fig.cap= "Association between car crashes and annaul average daily traffic volume", fig.width=4, fig.height=2}

# higher average daily traffic volume, linear positive increase in crashes
ggplot(crash, aes(y = log(crash + 1), x = adt)) +
  geom_point() +
  geom_smooth() +
  theme_bw() +
  theme(axis.title = element_text(face = "bold", size = 8)) +
  labs(y = "Log transformation of crash counts", x = "Annual average daily traffic volume in the unit vehicles per day" )

```

From the bi-variate plots, I then explored a select group of variables to be faceted by Botts' dots presence to investigate potential interactions between multiple co-variates and `crash` but nothing appeared too significant (**Appendix Part B**).


## Methods

**Model type**
\
The outcome of interest, `crash` counts, did not meet the assumption of a Gaussian distribution (**Fig. 1**) so a generalized linear model (GLM) was chosen for the analysis. The GLM has three components: random component (exponential family), systematic component (linear predictor: $\eta = \sum^p_{j=1} \beta_jx_j$), and link function (log link for Poisson). Because crashes were count data that were positively, right skewed and could take on positive values equal or greater than zero, a Poisson distribution was chosen as the special case of the exponential family which assumes the following density function

$$f(y;\theta, \phi) = exp(\frac{y\theta - b(\theta)}{a(\phi)} + c(y,\phi) )$$

where $\theta$ is the canonical parameter (= $ln\mu$ for Poisson) and $\phi$ is the dispersion parameter (= 1 for Poisson). The canonical link is $\eta = ln\mu$ and the dispersion parameter should equal or be close to 1. We assume that 1) y ~ $P(\mu)$, 2) the variance is V($\mu$) = $\mu$ and 3) the constraints on the mean are $\mu$ = E(y) > 0 so the ranges of $\mu$ and $\eta$ are different.
\
Although a Poisson model is a generalized model, the issue of over dispersion - where the variance of the response variable exceeds the nomial value, $Var(y) > \mu$ - is an issue that is common and needs to be addressed. To seek and model extra variation for over dispersion, a negative-binomial model and a quasi-likelihood model were explored with this dataset (**Appendix Part C**). A Poisson model with a offset was used to specify an a priori known component to be included in the linear predictor during fitting such as `adt` to eliminate over dispersion characteristics. Over dispersion was calculated for each model by dividing residual deviance by degree of freedom.

\
**Variable selection**
\
Multiple version of Poisson models (with offsets, negative binomial, quasipoisson) were attempted with single and multiple co-variates (**Appendix Part C**). Intuitively and based on exploratory data analysis, crash counts would be best explained by `adt` because the more vehicles that are on the road, they more accidents there are going to be (**Giarrantana 2019**) and for this specific report that is interested in the effectiveness of `botts` so both variables were important to include.
\
To decide additional variables that could best answer the specific question for this report there were two methods I went through *(1)* was a data driven approach where I ran exhaustive algorithms to calculate the AIC/BIC for models with all variables or a subset of them, and *(2)* a biologist intuition to build a model that made the most biological sense while adhering to general guidelines such as a parsimonious model that balances complexity with simplicity (ie. Occam's razor) (**Appendix Part D**). 

A null model (where, $\mu$ = constant) and saturated models (where, n parameters which leads to interpolation $\hat{\mu} = y_i$) were built. 

```{r, include = FALSE}

fit3 <- glm(crash ~ botts, family =  "poisson", data = crash)


fit16 <- glm(crash ~ botts + width + speed + shoulder + driveways + intersections + adt:speed + adt:driveways + offset(log(adt)), family = poisson, data = crash)

fit22 <- glm(crash ~ botts  + offset(log(adt)) + adt:speed + adt:driveways, data = crash, family = poisson)

summary(fit16)
summary(fit22)
```

\
**Model diagnostics**
\
Traditional model diagnostic tools such as qqPlots are not always as appropriate for GLMS. To evaluate departures from a model that could be either *(1)* systematic departure where an important covariate such as higher order terms or interactions could be missing or *(2)* isolated departure such as individual data points that contribute to consistency, leverage and influence of a model which is plotted in **Appendix Part B**.

\
**Model inference & estimation** 
\
The hypothesis we are testing for this report is if Botts' dots have an effect on counts of car crashes. The null hypothesis is that Botts' dots ($\beta$) have no effect on counts of car crashes ($H_0: \beta = 0$) while the alternative hypothesis is that Botts' dots do have an effect on counts of car crashes ($H_0: \beta \neq 0$) so the parameter value would be a non-0 integer. I tested this hypothesis with a chi-square test to indicate if `botts` are a statistically significant predictor of counts of car crashes (**Appendix Part D**). 
\
In addition to testing the effectiveness of Botts' dots I wanted to test for the best fit model with co-variates other than `botts`. I calculated the deviance and log-likelihood to provide a measure of goodness of fit as well as AIC/BIC scores to measure suitability and complexity for each of my models (**Appendix Part D**). 
\
After narrowing down to two candidate models, I used likelihood ratio with a Chi-square statistic to test if my more complex model (ie. **fit16**) with 9 parameters can be reduced to a sub-model (ie. **fit22**) with 4 parameters (**Appendix Part E**).

## Results

Counts of `car` crashes did not follow a Gaussian distribution (**Fig. 1**) so a GLM was chosen for the analysis. Because the response variable was count data, a Poisson model was the most appropriate however, over dispersion was a concern and the reason for attempting multiple model types (ie. Poisson, Quasi-poisson, Negative Binomial). To transform the non-linear relationship to linear form, a link function is used which is the log for Poisson regression. When the full model was attempted with each model type the Poisson model performed the best with the closest dispersion estimate to 1 (residual deviance/df = 1.08) so the results in this report will be based on a Poisson model but other model types attempts can be found in the **Appendix Part C**. 
\
To address the simple question, is there an effect of Botts' dots on car crashes a simple Poisson model was used.
\
\
**Fit3**$\Longrightarrow$ $$crash ~\sim botts$$
\
\
With this simple model (ie. **fit3**) there was a statistical significant difference between Botts' dot presence (= 1; **p = 0.00177**) and Botts' dots absence (= 0; **p < 2e-16**). The coefficient estimate for Botts' presence is -0.05026, meaning that the expected log count for a one-unit increase in Botts' dots presence is -0.05026 compared to Botts' absence. The coefficient estimate for Botts' absence is 0.46945. So if Botts' dots are not present, there is a positive association with car crashes and if Botts' are are present there is a negative association with car crashes, which is the intended purpose of Botts' dots (**Fig. 3**). 


```{r, fig.cap = "Fit 3 predictor effects of botts on car crashes",fig.width=5, fig.height=3, fig.align='center'}
# summary(fit3)
plot(predictorEffects(fit3),
     axes = list(y=list(type = "response", lab = "log(crash counts)"), 
                 x=list(botts = list(lab = "Botts present (0 = no, 1 = yes)"))))

```


However car crashes as the result of multiple factors, not just the presence of Botts' dots so additional models were carried out to evaluate the variance of car crashes within this dataset. For variable selection multiple models were attempted (n = 24) from a null model (ie. `car` ~ 1) to full model (`car` ~ all co-variates) and models with interactions which can be found in the **Appendix Part D**. However, the models that performed better (eg. based on AIC/BIC, $R^2$, p-value) included an offset option of `adt` which made biological sense as an exposure variable that impacts `car` and to improve computational speed, I logged transformed `adt`. Because the MS pass does not require to explore a trend in time, `year` was not included in the models although models that included an interaction between `year` and  `location` did perform better than those that did not have the interaction term (**Appendix Part C**). After multiple iterations of variable selection, there were two models that are strong candidates to explain the variance of car crashes.  


\

**Fit16**	$\Longrightarrow$ $$crash \sim botts + width + speed + shoulder + driveways + intersections + adt:speed + adt:driveways + offset(log(adt)) $$
```{r, include = FALSE}
#summary(fit16)
```

**Fit22** $\Longrightarrow$ $$crash \sim botts + offset(log(adt)) + adt:speed + adt:driveways$$
```{r, include = FALSE}
#summary(fit22)
```

\

When comparing the models with a likelihood ratio test (ie. Chi-square), the deviance for **fit16** is lower (13892) than **fit22** (13936) and there is a statistical difference between **fit16** and **fit22** (**p = 4.657e-07**). And although **fit16** performs slightly better (**AIC: 31046**, dispersion estimate: 1.391) than **fit22** (**AIC: 31074**, dispersion estimate: 1.394), I think **fit22** is the more preferred model because *(1)* it directly addresses the question of botts effects on car crashes with less co-variates and *(2)* it's a simpler model that performs somewhat similar to the more complex model of **fit16**. 

To summarize the output of **fit22**, `botts`, interaction terms `adt:speed` and `adt:driveways` had an effect on `crash`. `botts` was the most significant explanatory variable with a p-value of < 2e-16 for both presence & absence of `botts`. The next significant explanatory variables were interaction terms `adt:speedHigh` (**p = 0.000387**), `adt:driveways> 10` (**p = 0.007175**), and `adt:speedLow` (**p = 0.019085**) with `adt:driveways> 0 and <= 10` as a trending significant explanatory variable (**p = 0.067130**). The exponentiated coefficients estimates, standard error, and z-value can be found in **Fig. 4**. The model fit of **fit22**  had **AIC of 31074**, **BIC of  31117**, **pseudo-$R^2$ of 0.02**.

```{r fig.cap= "Fit22 summary with exponentiated coefficients"}

summ(fit22, confint = TRUE, digits = 3, exp = TRUE)

```


I also tested the overall effect of botts by comparing the deviance of the full model (ie. **fit22**, deviance = 13936) with the deviance of the model excluding botts (ie. **fit25**, deviance = 14062). The two degree-of-freedom chi-square test indicates that botts taken together is a statistically significant predictor of car crashes (**p < 2.2e-16**). 

```{r, include = FALSE}
#fit25 <- update(fit22, . ~ . - botts)

#anova(fit25, fit22, test = "Chisq")

```

## Discussion & Conclusion
\
Botts' dots do effect the count of car crashes based on this dataset and particular analyses. Roads that had Botts' dots were associated with lower counts of car crashes compared to roads without Botts' dots (**Fig. 3**). Whether it was Botts' dots in itself that reduced traffic accidents is difficult to fully evaluate but when comparing models that included `botts` versus not included `botts`, `botts` was a significant predictor of car crashes (chi-square; **p < 2.2e-16**). However it was important to include the offset of `adt` because of the strong correlation between average traffic volume and car crashes (**Fig. 2**) (**Olaisen 2019**). 

```{r, fig.cap = "Fit22 predictor effects on car crashes", fig.width=10, fig.height= 12}
par(mfrow = c(1,1))
plot(predictorEffects(fit22),
     axes = list(y=list(type = "response", lab = "log(crash counts)")))
```


In this dataset, additional co-variates were associated with car crashes and Botts' dots presence. The interaction between average daily traffic volume with other variables such as speed and driveways were significantly associated with car crash counts (**Fig. 5**). This may be due to the fact that more cars on the roads leads to more opportunities for cars to crash and traffic volume can impact speed, especially if there is traffic congestion. The interaction between `adt` and high `speed` was strongly significant with crashes (**fit22; p-value = 0.000387**) probably because at higher speeds people have less time to react to things that may cause a car crash. Another interaction term that was significant but not quite as clear of biological association is `adt`and 10+ `driveways` (**fit22; 0.007175**). I am assuming places with lots of driveways are maybe in a more urban or sub-urban setting compared to rural counterparts. Urban settings tend to have higher population densities which may have more traffic volume and more obstacles that drivers have to overcome to avoid an accident.

Driving is a higher-order thinking skill that can be impacted by many things that are not infrastructure related such as age and gender (**Spencer 2021**) or anomaly weather conditions such as flooding, high speed winds, or dense fog. A dataset that included some of those biotic and abiotic factors would be helpful to determine if Botts' dots are truly a significant factor in car crashes or just a result of a dataset that lacks complexity. Also there are many trends in this dataset, especially with regards to width of shoulder or length of road, that I may not fully understand their relevance due to my lack of infrastructure knowledge such as a city planner or engineer may have. I would consult those experts before making a final recommendation of how to make roads safer for all drivers. 

## Appendix
An appendix of code for normality check, exploratory data analysis, model fitting, variable selection, and model selection can be found in the attached document labeled "PSTAT_2022QualExam_MS_Sambado_Appendix.pdf".


## References

**Olaisen** RH, Rossen LM, Warner M, Anderson RN. Unintentional injury death rates in rural and urban areas: United States, 1999–2017pdf icon. NCHS Data Brief, no 343. Hyattsville, MD: National Center for Health Statistics. 2019.

**Spencer** MR, Hedegaard H, Garnett M. Motor vehicle traffic death rates, by sex, age group, and road user type: United States, 1999–2019. NCHS Data Brief, no 400. Hyattsville, MD: National Center for Health Statistics. 2021. DOI: https://dx.doi.org/10.15620/cdc:101759external icon

**Giarrantana C** https://www.trafficsafetystore.com/blog/4-ways-cities-are-using-smart-technology-to-control-traffic-congestion/

**Wikipedia** https://en.wikipedia.org/wiki/Botts%27_dots