---
title: 'EEMB 179 -- Lab 5: Managing the recovery of predator-prey systems'
---

Announcements --   
  (1) Paper selection for final project should be finalized by now. 
  (2) **NEW DEADLINE**: Thursday, February 27th by 11:59pm -- Submit code for your model on Gauchospace (for homework points). For full credit, show us an attempt at a timeseries (N vs. t) and a written interpretation of the graph. This exercise functions to encourage you to start working on your code. Your TAs will *not* diagnose the code you submit if it does not work, but are happy to meet with you during office hours, appointments or the free lab period (Week 9).
  (3) Grades for the Midterm have been posted on Gauchospace. Please double-check your scores were graded, added up and entered correctly. See Dr. Moeller's email for info regarding the answer key and regrade policy. 
  
This lab serves two purposes:
  (1) Provide a demonstration of the expectations for your final project, and
  (2) Complement what we're learning in lecture about predator-prey dynamics.

Before coming to lab, you should carefully read "Rapid and direct recoveries of predators and prey through synchronized ecosystem management" by Samhouri et al. (2017). Dr. Moeller will start lecture by presenting a 15-minute "chalk talk" about the paper. We will then work with code to recreate the paper's predictions and test the effects of some of the authors' core assumptions.

# PART ONE: Studying the unharvested system

The equations for the model are given in Box 1, but the parameter values used for the simulations are hidden away in the tiny font Methods section that always appears at the end of papers in the Nature series of journals. 

The model has two species whose populations change in time: X, a prey species, and P, a predator. Although there is a second prey species "Y," this species's abundance is held constant over time, so for the purposes of the paper, Y is a parameter, not a state variable.

The biological dynamics (ignoring harvest for now) of each species are given by:

$$
\begin{align}
\frac{d X}{dt}  &= r_x X (1-\frac{X}{K_x}) - a_x P X \\ 
\newline
\frac{d P}{dt} &= P [c(a_x X + a_y Y) - d_p] (1-\frac{P}{K_p}) \\
\newline
\end{align}
$$

# PART 1A:
# Think, Pair, Share 1 

1. Make a list of all the variables in the model and their meanings

 X = density of prey
 P = density of predator
 t = time

2. Make a list of all the parameters in the model and their meanings.
 
 r_x = growth rate of the prey
 K_x = carrying capacity of the prey
 a_x = attack rate of the predator on prey X
 c = conversion efficiency of both prey types into predator biomass
 a_y = attack rate of the predator on prey Y
 d_p = mortality rate of the predator
 K_p = carrying capacity of the predator
 Y = number of the other prey type

3. What type of growth do the prey and predator exhibit?

 Logistic

# PART 1B: Let's set the values of each of the parameters 
```{r}
r_x <- 1 		# prey X's growth rate
K_x <- 100 		# prey X's carrying capacity
a_x <- 0.03		# attack rate of predator on prey X
c <- 0.05		# predator's conversion efficiency
a_y <- 0.03		# attack rate of predator on prey Y
d_p <- 0.25		# mortality rate of predator
K_p <- 25		# predator's carrying capacity
Y <- 500		# abundance of other prey type
```

# PART 1C: And choose some initial conditions
```{r}
X0 <- 1 # prey X initial population size
P0 <- 1 # predator initial population size
```

# PART 1D: Now, we're ready to simulate!
```{r}
tset <- seq(0,100,length.out=20000)
X.simu1 <- NaN*tset
X.simu1[1] <- X0
P.simu1 <- NaN*tset
P.simu1[1] <- P0

for(i in 2:length(tset)){
	dt <- tset[i]-tset[i-1]
	X <- X.simu1[i-1]
	P <- P.simu1[i-1]
	dX <- ( r_x*X*(1-X/K_x) - a_x*P*X ) * dt
	dP <- ( P*(c*(a_x*X+a_y*Y)-d_p)*(1-P/K_p) ) * dt
	X.simu1[i] <- X.simu1[i-1] + dX
	P.simu1[i] <- P.simu1[i-1] + dP
}
```

# PART 1E: And, time to plot!
```{r}
Xcol <- 'deepskyblue'
Pcol <- 'firebrick'

plot(tset,X.simu1,type='l',xlab='Time',ylab='Density',las=1,lwd=2,col=Xcol)
lines(tset,P.simu1,lwd=2,col=Pcol)
legend(x = 70, y = 50, legend = c('Prey','Predator'),lwd=2,col=c(Xcol,Pcol))
```

# PART TWO: Exploiting the system

# PART 2A: Setting new initial conditions

Samhouri et al. consider the unexploited system (i.e., with no hunting) to be the 'baseline' case. They then allow for exploitation by altering the equations to add a harvest term to each:

$$
\begin{align}
\frac{d X}{dt}  &= r_x X (1-\frac{ X}{K_x}) - a_x P X - h_x X\\ 
\newline
\frac{d P}{dt} &= P [c(a_x X + a_y Y) - d_p] (1-\frac{P}{K_p}) - h_p P\\
\newline
\end{align}
$$

So, let's set our initial conditions to be the equilibrium state of the system without hunting:
```{r}
X0 <- X.simu1[length(tset)]
P0 <- P.simu1[length(tset)]
```

# PART 2B: Getting ready to hunt!

We will also need to set two new parameters for the hunting rates
```{r}
h_x <- 0.65 		# hunting rate of species X
h_p <- 0.325		# hunting rate of species P
```

If we look at Samhouri et al. Figure 2, panel a, we can see that during the "disassembly" of the system, predator hunting is initiated first (around t = 30), and prey hunting is started later (around t = 60). We can save these timepoints by assigning them to variables.
```{r}
t_Pstart <- 30
t_Xstart <- 60
```

# PART 2C: Simulating hunting with a for loop

In order to recreate Figure 2a, we are going to have to complicate our for loop a bit. In particular, we need to be able to turn hunting "on" at the timepoints chosen above. In order to do this, we'll insert "if" statements into our for loop that allow for hunting only when the time is later than the startpoints chosen in the section above.
```{r}
tset <- seq(0,100,length.out=20000)
X.simu2 <- NaN*tset
X.simu2[1] <- X0
P.simu2 <- NaN*tset
P.simu2[1] <- P0

for(i in 2:length(tset)){
	
	# h_p_now and h_x_now are "dummy variables" that are equal to 0 when hunting isn't happening, and equal to h_p and h_x respectively when hunting is happening.
	
	# The If/Else statements below say, "If this time is before the start time of hunting, then hunting = 0. Otherwise, hunting = the fixed hunting rate for that species."
	
	if(tset[i] < t_Pstart){ 	h_p_now <- 0 } else { h_p_now <- h_p }
	
	if(tset[i] < t_Xstart){ 	h_x_now <- 0 } else { h_x_now <- h_x }
	
	dt <- tset[i]-tset[i-1]
	X <- X.simu2[i-1]
	P <- P.simu2[i-1]
	dX <- ( r_x*X*(1-X/K_x) - a_x*P*X - h_x_now*X ) * dt
	dP <- ( P*(c*(a_x*X+a_y*Y)-d_p)*(1-P/K_p) - h_p_now*P ) * dt
	X.simu2[i] <- X.simu2[i-1] + dX
	P.simu2[i] <- P.simu2[i-1] + dP
}
```

# PART 2D: Plotting results. Cross your fingers that it looks like Fig. 2a!
```{r}
plot(tset,X.simu2,type='l',xlab='Time',ylab='Density',las=1,lwd=2,col=Xcol,ylim=c(0,65))
lines(tset,P.simu2,lwd=2,col=Pcol)
legend(x = 70, y = 65, legend = c('Prey','Predator'),lwd = 2,col=c(Xcol,Pcol))
# We can add vertical lines that indicate where hunting was turned on for each species.
abline(v = t_Pstart, lty = 2, col = Pcol)
abline(v = t_Xstart, lty = 2, col = Xcol)
```

# PART THREE: Recovering the system

# PART 3A: Setting new initial conditions and time windows

Samhouri et al. consider three 'recovery' scenarios: One in which hunting of prey and predator is stopped simultaneously, one in which hunting of the predator is stopped first, and one in which hunting of the prey is stopped first.

Because the system is going to be recovering, our initial conditions need to be those of the "disassembled" state (which we just plotted).
```{r}
X0_dis <- X.simu2[length(tset)]
P0_dis <- P.simu2[length(tset)]
```

We're also going to be running our simulation over a different range of time, picking up where we left off above at t = 100, and running the model for a further 110 years. Note that I've set the length of the tset2 vector to 21,000 (compared to 20,000 for tset). This is a coding 'trick' to help me catch errors when I make plots later. If I try to plot one of my model results (a .simu vector) with the wrong timeset, I'll get an error message telling me that the dimensions of the vectors don't agree.
```{r}
tset2 <- seq(100,210,length.out = 21000)
```

# PART 3B: THINK, PAIR, SHARE 2
1. Study Samhouri et al. Figure 2, panel b. At what time are prey and predator hunting stopped? Assign this time to a parameter called t_stophunt.
```{r}
### solution ###

t_stophunt <- 120

### end solution ###
```

2. From Figure 2, panel c, at what time is the prey hunting stopped? Assign this time to a parameter called t_stopX.
```{r}
### solution ###

t_stopX <- 170
t_stopX <- 140
### end solution ###
```

3. From Figure 2, panel d, at what time is the predator hunting stopped? Assign this time to a parameter called t_stopP.
```{r}
### solution ###

t_stopP <- 150

### end solution ###
```

# PART 3C: Scenario 1: Synchronous recovery

Now, we're ready to run our first simulation! Pay close attention to how the 'if' statements in the for loop have been modified. This should recreate Fig 2b. 
```{r}
X.simu3 <- NaN*tset2
X.simu3[1] <- X0_dis
P.simu3 <- NaN*tset2
P.simu3[1] <- P0_dis

for(i in 2:length(tset2)){
	
	if(tset2[i] < t_stophunt){ h_p_now <- h_p } else { h_p_now <- 0 }
	
	if(tset2[i] < t_stophunt){ h_x_now <- h_x } else { h_x_now <- 0 }
	
	dt <- tset2[i]-tset2[i-1]
	X <- X.simu3[i-1]
	P <- P.simu3[i-1]
	dX <- ( r_x*X*(1-X/K_x) - a_x*P*X - h_x_now*X ) * dt
	dP <- ( P*(c*(a_x*X+a_y*Y)-d_p)*(1-P/K_p) - h_p_now*P ) * dt
	X.simu3[i] <- X.simu3[i-1] + dX
	P.simu3[i] <- P.simu3[i-1] + dP
}

plot(tset2,X.simu3,type='l',xlab='Time',ylab='Density',las=1,lwd=2,col=Xcol,ylim=c(0,65))
lines(tset2,P.simu3,lwd=2,col=Pcol)
legend(x = 170, y = 65, legend = c('Prey','Predator'),lwd = 2,col=c(Xcol,Pcol))
abline(v = t_stophunt, lty = 2, col = Pcol)
abline(v = t_stophunt, lty = 3, col = Xcol)
```

# PART 3D: Scenario 2: Predator-first recovery

Now we'll run the second possibility and try to re-create the graph in Figure 2c. Note that we're saving this as our fourth simulation (X.simu4 and P.simu4)! 
```{r}
X.simu4 <- NaN*tset2
X.simu4[1] <- X0_dis
P.simu4 <- NaN*tset2
P.simu4[1] <- P0_dis

for(i in 2:length(tset2)){
	
	if(tset2[i] < t_stophunt){ h_p_now <- h_p } else { h_p_now <- 0 }
	
	if(tset2[i] < t_stopX){ h_x_now <- h_x } else { h_x_now <- 0 }
	
	dt <- tset2[i]-tset2[i-1]
	X <- X.simu4[i-1]
	P <- P.simu4[i-1]
	dX <- ( r_x*X*(1-X/K_x) - a_x*P*X - h_x_now*X ) * dt
	dP <- ( P*(c*(a_x*X+a_y*Y)-d_p)*(1-P/K_p) - h_p_now*P ) * dt
	X.simu4[i] <- X.simu4[i-1] + dX
	P.simu4[i] <- P.simu4[i-1] + dP
}
```

# THINK, PAIR, SHARE 3: 
1. Using the plotting code from PART 3C above, make a plot of these results.
```{r}
### solution ###

plot(tset2,X.simu4,type='l',xlab='Time',ylab='Density',las=1,lwd=2,col=Xcol,ylim=c(0,65))
lines(tset2,P.simu4,lwd=2,col=Pcol)
legend(x = 170, y = 65, legend = c('Prey','Predator'),lwd = 2,col=c(Xcol,Pcol))
abline(v = t_stophunt, lty = 2, col = Pcol)
abline(v = t_stopX, lty = 3, col = Xcol)

### end solution ###
X.simu4[length(tset2)]

X.simu4[length(tset2)*.5]
```

2. Did you get the same results as Fig. 2c?


3. Try lowering t_stopX (for example, to 140) and re-run the code above. (Don't forget to change where the vertical line is drawn, too!) What happened?

```{r}
# Students can either lower t_stopX or increase tset2 to get a figure that looks like 2c. 
```

# PART 3E: Scenario 3: Prey-first recovery

# THINK, PAIR, SHARE 4: As you have for Scenarios 1 and 2, simulate recovery, but, in this case, consider the prey-first recovery. Hint: Copy/paste code from above, but don't forget to create new storage vectors (e.g., X.simu5 and P.simu5) to hold your results!
```{r}
### solution ###

X.simu5 <- NaN*tset2
X.simu5[1] <- X0_dis
P.simu5 <- NaN*tset2
P.simu5[1] <- P0_dis

for(i in 2:length(tset2)){
	
	if(tset2[i] < t_stopP){ h_p_now <- h_p } else { h_p_now <- 0 }
	
	if(tset2[i] < t_stophunt){ h_x_now <- h_x } else { h_x_now <- 0 }
	
	dt <- tset2[i]-tset2[i-1]
	X <- X.simu5[i-1]
	P <- P.simu5[i-1]
	dX <- ( r_x*X*(1-X/K_x) - a_x*P*X - h_x_now*X ) * dt
	dP <- ( P*(c*(a_x*X+a_y*Y)-d_p)*(1-P/K_p) - h_p_now*P ) * dt
	X.simu5[i] <- X.simu5[i-1] + dX
	P.simu5[i] <- P.simu5[i-1] + dP
}

plot(tset2,X.simu5,type='l',xlab='Time',ylab='Density',las=1,lwd=2,col=Xcol,ylim=c(0,65))
lines(tset2,P.simu5,lwd=2,col=Pcol)
legend(x = 170, y = 65, legend = c('Prey','Predator'),lwd = 2,col=c(Xcol,Pcol))
abline(v = t_stopP, lty = 2, col = Pcol)
abline(v = t_stophunt, lty = 3, col = Xcol)

### end solution ###
```

# PART FOUR: Lining up the scenarios in time

For fun, and to build our plotting skills, let's plot the "whole scenario," starting from an unharvested ecosystem, showing exploitation, followed by **synchronous** recovery.
```{r}
plot(tset,X.simu2,type='l',xlab='Time',ylab='Density',las=1,lwd=2,col=Xcol,ylim=c(0,65),xlim=c(0,210))
lines(tset2,X.simu3,lwd=2,col=Xcol)
lines(tset,P.simu2,lwd=2,col=Pcol)
lines(tset2,P.simu3,lwd=2,col=Pcol)
legend(x = 150, y = 65, legend = c('Prey','Predator'),lwd = 2,col=c(Xcol,Pcol))
abline(v=t_Xstart,lty=2,col=Xcol)
abline(v=t_Pstart,lty=2,col=Pcol)
abline(v = t_stophunt, lty = 2, col = Pcol)
abline(v = t_stophunt, lty = 3, col = Xcol)
```

Those vertical lines are a little cumbersome. Let's borrow a page from the paper and use coloured bars to indicate when each species is being harvested instead.
```{r}
plot(tset,X.simu2,type='l',xlab='Time',ylab='Density',las=1,lwd=2,col=Xcol,ylim=c(0,65),xlim=c(0,210))
lines(tset2,X.simu3,lwd=2,col=Xcol)
lines(tset,P.simu2,lwd=2,col=Pcol)
lines(tset2,P.simu3,lwd=2,col=Pcol)
legend(x = 150, y = 65, legend = c('Prey','Predator'),lwd = 2,col=c(Xcol,Pcol))
lines(x=c(t_Xstart,t_stophunt),y=c(3,3),lwd=10,col=Xcol)
lines(x=c(t_Pstart,t_stophunt),y=c(0,0),lwd=10,col=Pcol)
text(x = 142, y = 2.8, 'X being hunted',cex=.7)
text(x = 142, y = -.2, 'P being hunted',cex=.7)
```

# PART FIVE: Using bifurcation diagrams to evaluate model assumptions

One of the major assumptions that has always bothered me about this paper is that there is a second prey species, Y, which is not dynamic. In other words, it's not being affected by hunting, nor is it being affected by the preator population. 

Because Y is not dynamic, it's a parameter in this system. So we can make a bifurcation diagram studying the equilibria of the model with respect to Y.

There are two ways we can do this: Analytically (on pen and paper, using algebra to solve dP/dt = dX/dt = 0), or numerically. Mathematical ecologists will usually take an analytical approach whenever possible because (1) it's more accurate because you also find the unstable equilibria that your simulations will not reveal, (2) it's less computationally intensive, and (3) it's more fun! But for some of your projects, you'll need to generate bifurcation diagrams numerically, so that's what we'll do here. 

Because numerical approaches are bad at finding unstable equilibria (because there's almost no chance you'll land, unperturbed, on that equilibrium and stay there over simulation time), we'll ignore the unstable equilibrium at X = 0, P = 0, though this equilibrium is readily apparently from the equations. (Think about how you could factor out an X from dX/dt and a P from dP/dt, and what that means when you set those equations equal to zero to find equilibria.)

# PART 5A: Selecting the x-axis for our bifurcation diagram

We'll use 'seq' to create a sequence of values of Y for which we'll look at the model's equilibria.
```{r}
Yset <- seq(0,1000,length.out = 50)
```

# PART 5B: Calculating the equilibrium points for each value of Y

We need to do a loop of loops: for each value of Y within Yset, we'll run the numerical simulation (looping over all values of t in tset), and save the output as the equilibrium.

NOTE: For this to work, you have to be confident that your timeset is long enough for the model to reach equilibrium! Based on our simulations in Part One, we can feel pretty good about having time run to 100:
```{r}
plot(tset,X.simu1,type='l',xlab='Time',ylab='Density',las=1,lwd=2,col=Xcol)
lines(tset,P.simu1,lwd=2,col=Pcol)
legend(x = 70, y = 50, legend = c('Prey','Predator'),lwd = 2,col=c(Xcol,Pcol))
```

First, we create storage variables for our two state variables of interest. 
```{r}
Xstarset <- NaN*Yset
Pstarset <- NaN*Yset
```

Second, we run our for loop. Note that I'm using 'j' as my index of Y values. It's important that each nested for loop using a different index variable. Otherwise, indices are getting reset within loops and you could get stuck in a loop of loops of loops of....
```{r}
for(j in 1:length(Yset)){
	Y <- Yset[j] 
	
	X.simu <- NaN*tset # create a holding vector for the prey population X of the same length as tset (20,000)
	X.simu[1] <- 1   
	P.simu <- NaN*tset 
	P.simu[1] <- 1
	
	for(i in 2:length(tset)){
		dt <- tset[i]-tset[i-1]
		X <- X.simu[i-1]
		P <- P.simu[i-1]
		dX <- ( r_x*X*(1-X/K_x) - a_x*P*X - h_x*X ) * dt
		dP <- ( P*(c*(a_x*X+a_y*Y)-d_p)*(1-P/K_p) - h_p*P ) * dt
		X.simu[i] <- X.simu[i-1] + dX
		P.simu[i] <- P.simu[i-1] + dP	
	}
	
	Xstarset[j] <- X.simu[length(tset)]
	Pstarset[j] <- P.simu[length(tset)]
	
}
```

# PART 5C: Examining the results

# We can now make bifurcation diagrams for both X and P
```{r}
plot(Yset,Pstarset,type='l',lwd=2,col=Pcol,las=1,xlab='Density of alternate prey, Y',ylab='Stable equilibria of predator, P*')

plot(Yset,Xstarset,type='l',lwd=2,col=Xcol,las=1,xlab='Density of alternate prey, Y',ylab='Stable equilibria of focal prey, X*')
```

# PART SIX: HOMEWORK

```{r}
# 1. Make a timeseries plot like that in Part 4 (i.e., showing a timeseries from year 0 to year 210) for the scenario in which prey are allowed to recover before predators (i.e., combining figures 2a and 2d from Samhouri et al.). Be sure to (1) label your axes, (2) include a legend, and (3) indicate when harvest of predators and prey is occurring.a

plot(tset,X.simu2,type='l',xlab='Time',ylab='Density',las=1,lwd=2,col=Xcol,ylim=c(0,68),xlim=c(0,230))
lines(tset2,X.simu5,lwd=2,col=Xcol)
lines(tset,P.simu2,lwd=2,col=Pcol)
lines(tset2,P.simu5,lwd=2,col=Pcol)
legend(x = 158, y = 65, legend = c('Prey','Predator'),lwd = 2,col=c(Xcol,Pcol))
lines(x=c(t_Xstart,t_stophunt),y=c(3.9,3.9),lwd=10,col=Xcol)
lines(x=c(t_Pstart,t_stopP),y=c(0,0),lwd=10,col=Pcol)
text(x = 147, y = 4, 'X being hunted',cex=.7)
text(x = 178, y = 0, 'P being hunted',cex=.7)
```
/1 axes
/1 legend
/1 graphical indication of predator harvest
/1 graphical indication of prey harvest
/2 one point per simulation line (exploitation phase) 
/2 one point per simulation line (recovery phase)
= /8 points total

```{r}
# 2. To interpret the effects of changing alternate prey availability -- 
# 2a. Paste your bifurcation diagrams from Part 5C here. (You should have two -- one for X*, and one for P*, both with respect to Y.) 

plot(Yset,Pstarset,type='l',lwd=2,col=Pcol,las=1,xlab='Density of alternate prey, Y',ylab='Stable equilibria of predator, P*')

plot(Yset,Xstarset,type='l',lwd=2,col=Xcol,las=1,xlab='Density of alternate prey, Y',ylab='Stable equilibria of focal prey, X*')
```
/2 axes for two plots
/2 X* or P* lines in two plots
= /4 points total

```{r}
# 2b. Describe in biological terms what is happening as the alternate prey increases in density.
# The density of predators at equilibrium (P*) increases as a function of alternate prey density, but only past a threshold of alternate prey subsidy (~350 alternate prey). Conversely, the focal prey at equilibrium (X*) decreases as a function of alternative prey density, but only past the same threshold (~350) of alternate prey where predator population size increases. 
```
= /2 points total

```{r}
# 2c. Look up the term "apparent competition." How does your bifurcation diagram of X* with respect to Y show this phenomenon?

# Apparent competition -- A form of competition between species or group of organisms indirectly competing with another species or group of organisms, which both of them serve as prey of a predator. The bifurcation diagram of X* with respect to Y shows this phenomenon because increases in alternate prey density causes an indirect and negative effect on the focal prey species by increasing the predator population. 
```
= /2 points total

```{r}
# 3. Choose your own (bifurcation) adventure.
# 3a. Choose a parameter other than Y, and make two bifurcation diagrams with this parameter on the x-axis. One bifurcation diagram should show the stable prey equilibrium (X*), and the other should show the stable predator equilibrium (P*).
h_xset <- seq(0,1,length.out=50)
Xstarset <- NaN*h_xset
Pstarset <- NaN*h_xset
for(j in 1:length(h_xset)){
	h_x <- h_xset[j] 
	
	X.simu2 <- NaN*tset # create a holding vector for the prey population X of the same length as tset (20,000)
	X.simu2[1] <- 1   
	P.simu2 <- NaN*tset 
	P.simu2[1] <- 1
	
	for(i in 2:length(tset2)){
		dt <- tset2[i]-tset2[i-1]
		X <- X.simu2[i-1]
		P <- P.simu2[i-1]
		dX <- ( r_x*X*(1-X/K_x) - a_x*P*X - h_x*X ) * dt #used a different varible other than i
		dP <- ( P*(c*(a_x*X+a_y*Y)-d_p)*(1-P/K_p) - h_p*P ) * dt
		X.simu2[i] <- X.simu2[i-1] + dX
		P.simu2[i] <- P.simu2[i-1] + dP	
	}
	
	Xstarset[j] <- X.simu2[length(tset2)] #these are holding vectors for equilibrium values
	Pstarset[j] <- P.simu2[length(tset2)]
	
}
plot(h_xset,Pstarset,type='l',lwd=2,col=Pcol,las=1,xlab='Prey Hunting, H',ylab='Stable equilibria of predator, P*')

plot(h_xset,Xstarset,type='l',lwd=2,col=Xcol,las=1,xlab='Prey Hunting, H',ylab='Stable equilibria of focal prey, X*')


```
/2 axes for two plots
/2 X* or P* lines in two plots
= /4 points total

```{r}
# 3b. Describe in biological terms what is happening. Do your results make intuitive sense?
```
= /2 points total

```{r}
# 4. Study the impacts of reduced alternate prey on recovery scenarios by setting Y = 200.
# 4a. Using the same harvest rates we have chosen above (h_x = 0.65 and h_p = 0.325), find the equilibrium population sizes when harvest is occurring (i.e., in a "disassembled" or "exploited" system). Set h_x and h_p to be ON the entire simulation. 
```
/1 equilibrium value for prey
/1 equilibrium value for predator
/2 code and graph that show work by simulation
= /4 points total

```{r}
### solution ###
r_x <- 1 		# prey X's growth rate
K_x <- 100 		# prey X's carrying capacity
a_x <- 0.03		# attack rate of predator on prey X
c <- 0.05		# predator's conversion efficiency
a_y <- 0.03		# attack rate of predator on prey Y
d_p <- 0.25		# mortality rate of predator
K_p <- 25		# predator's carrying capacity
h_x = 0.65 
h_p = 0.325

Y <- 200

X.simu10 <- NaN*tset
P.simu10 <- NaN*tset
X.simu10[1] <- 1
P.simu10[1] <- 1

for(i in 2:length(tset)){
	dt <- tset[i]-tset[i-1]
	X <- X.simu10[i-1]
	P <- P.simu10[i-1]
	dX <- ( r_x*X*(1-X/K_x) - a_x*P*X - h_x*X ) * dt
	dP <- ( P*(c*(a_x*X+a_y*Y)-d_p)*(1-P/K_p) - h_p*P ) * dt
	X.simu10[i] <- X.simu10[i-1] + dX
	P.simu10[i] <- P.simu10[i-1] + dP	
}

X.simu10[length(tset)]
P.simu10[length(tset)]

# NOTE: figure not necessary for full credit! I just have it in here to check that I'm at an equilibrium #
plot(tset,X.simu10,las=1,type='l',lwd=2,col=Xcol,xlab='Time',ylab='Density',ylim=c(0,65)); lines(tset,P.simu10,lwd=2,col=Pcol); legend(x = 70, y = 65, legend = c('Prey','Predator'),lwd = 2,col=c(Xcol,Pcol))

### end solution ###
```

```{r}
# 4b. Simulate the three recovery scenarios (synchronous, predator-first, and prey-first). Create three timeseries graphs showing your results. Be sure to label your axes, include a legend, and indicate the time at which harvest stopped. HINT: You may need to run your timeseries for longer! Try tset3 <- seq(100,500,length.out=22000)
```
/3 axes for three plots
/6 predator and prey lines for three plots
/6 graphical indication of when hunting stops
/3 legends for three plots
= /18 points total

```{r}
### solution ###

X0_alt <- X.simu10[length(tset)]
P0_alt <- P.simu10[length(tset)]

# Simultaneous
tset3 <- seq(100,500,length.out=22000)
X.simu11 <- NaN*tset3
X.simu11[1] <- X0_alt
P.simu11 <- NaN*tset3
P.simu11[1] <- P0_alt

for(i in 2:length(tset3)){
	if(tset3[i] < t_stophunt){ h_p_now <- h_p } else { h_p_now <- 0 }
	if(tset3[i] < t_stophunt){ h_x_now <- h_x } else { h_x_now <- 0 }
	dt <- tset3[i]-tset3[i-1]
	X <- X.simu11[i-1]
	P <- P.simu11[i-1]
	dX <- ( r_x*X*(1-X/K_x) - a_x*P*X - h_x_now*X ) * dt
	dP <- ( P*(c*(a_x*X+a_y*Y)-d_p)*(1-P/K_p) - h_p_now*P ) * dt
	X.simu11[i] <- X.simu11[i-1] + dX
	P.simu11[i] <- P.simu11[i-1] + dP
}

plot(tset3,X.simu11,type='l',xlab='Time',ylab='Density',las=1,lwd=2,col=Xcol,ylim=c(0,100))
lines(tset3,P.simu11,lwd=2,col=Pcol)
legend(x = 350, y = 65, legend = c('Prey','Predator'),lwd = 2,col=c(Xcol,Pcol))
abline(v = t_stophunt, lty = 2, col = Pcol)
abline(v = t_stophunt, lty = 3, col = Xcol)

# Pred-first
X.simu12 <- NaN*tset3
X.simu12[1] <- X0_alt
P.simu12 <- NaN*tset3
P.simu12[1] <- P0_alt
for(i in 2:length(tset3)){
	if(tset3[i] < t_stophunt){ h_p_now <- h_p } else { h_p_now <- 0 }
	if(tset3[i] < t_stopX){ h_x_now <- h_x } else { h_x_now <- 0 }
	dt <- tset3[i]-tset3[i-1]
	X <- X.simu12[i-1]
	P <- P.simu12[i-1]
	dX <- ( r_x*X*(1-X/K_x) - a_x*P*X - h_x_now*X ) * dt
	dP <- ( P*(c*(a_x*X+a_y*Y)-d_p)*(1-P/K_p) - h_p_now*P ) * dt
	X.simu12[i] <- X.simu12[i-1] + dX
	P.simu12[i] <- P.simu12[i-1] + dP
}

plot(tset3,X.simu12,type='l',xlab='Time',ylab='Density',las=1,lwd=2,col=Xcol,ylim=c(0,100))
lines(tset3,P.simu12,lwd=2,col=Pcol)
legend(x = 350, y = 65, legend = c('Prey','Predator'),lwd = 2,col=c(Xcol,Pcol))
abline(v = t_stophunt, lty = 2, col = Pcol)
abline(v = t_stopX, lty = 3, col = Xcol)


# Prey-first
X.simu13 <- NaN*tset3
X.simu13[1] <- X0_alt
P.simu13 <- NaN*tset3
P.simu13[1] <- P0_alt
for(i in 2:length(tset3)){
  	if(tset3[i] < t_stopP){ h_p_now <- h_p } else { h_p_now <- 0 }
	if(tset3[i] < t_stophunt){ h_x_now <- h_x } else { h_x_now <- 0 }

	dt <- tset3[i]-tset3[i-1]
	X <- X.simu13[i-1]
	P <- P.simu13[i-1]
	dX <- ( r_x*X*(1-X/K_x) - a_x*P*X - h_x_now*X ) * dt
	dP <- ( P*(c*(a_x*X+a_y*Y)-d_p)*(1-P/K_p) - h_p_now*P ) * dt
	X.simu13[i] <- X.simu13[i-1] + dX
	P.simu13[i] <- P.simu13[i-1] + dP
}

plot(tset3,X.simu13,type='l',xlab='Time',ylab='Density',las=1,lwd=2,col=Xcol,ylim=c(0,100))
lines(tset3,P.simu13,lwd=2,col=Pcol)
legend(x = 350, y = 65, legend = c('Prey','Predator'),lwd = 2,col=c(Xcol,Pcol))
abline(v = t_stopP, lty = 2, col = Pcol)
abline(v = t_stophunt, lty = 3, col = Xcol)



### end solution ###
```

```{r}
# 4c. Compare your findings to one another. For which scenario was overall recovery to an unexploited state fastest?
```
= /2 points total

```{r}
### solution ###

# Note: These overlay plots aren't necessary, but they are helping me visualize the correct answer, which is that simultaneous recovery gets you to the unexploited state fastest. Prey-first is actually slowest (because the rate limiting step seems to be predator recovery from very low population sizes, so the earlier you stop harvesting them, the better).

plot(tset3,X.simu11,type='l',xlab='Time',ylab='Density',las=1,lwd=2,col=Xcol,ylim=c(0,100))
lines(tset3,X.simu12,lwd=2,col='deepskyblue4')
lines(tset3,X.simu13,lwd=2,col='deepskyblue3')
legend(x = 350, y = 80, legend=c('Simultaneous','Pred-first','Prey-first'),lwd=2,col=c(Xcol,'deepskyblue4','deepskyblue3'))


plot(tset3,P.simu11,type='l',xlab='Time',ylab='Density',las=1,lwd=2,col=Pcol,ylim=c(0,25))
lines(tset3,P.simu12,lwd=2,col='firebrick4')
lines(tset3,P.simu13,lwd=2,col='firebrick1')
legend(x = 350, y = 20, legend=c('Simultaneous','Pred-first','Prey-first'),lwd=2,col=c(Pcol,'firebrick4','firebrick1'))

### end solution ###
```

```{r}
# 4d. Compare your findings to those of Samhouri et al. (2017). What do your results suggest about the importance of a secondary prey type to predator-first recovery?

### solution ###

# Student should say something about a secondary prey type being necessary for the predator to mount some kind of recovery on its own (e.g., in a predator-first scenario). That secondary prey type is needed to support the predator's 'comeback,' so that in turn the predator can supress the population explosion of the secondary prey type when hunting on that species is halted.


### end solution ###
```
= /2 points total

= /48 points total







